---
title: | 
      | Practical Data Wrangling With R
      | Basics
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: zenburn
    css: documentCSS.css
---
```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = FALSE, comment = "")
```

# Symbols {.tabset .tabset-fade .tabset-pills}

Here are some basic symbols that you will see in R:

## <-

An assignment operator
  
```{r}
x <- 1:5
```
  
## =   
  
An alternative assignment operator (not preferred by the purists)
  
```{r}
y = 6:10
```
  
## ()  
  
() will always be for a function -- for example mean() 

```{r}
mean(x)
```
  
## \{\}  
  
{} always contain calls to created functions

```{r}
function(x) {
  x + 5
}
```

## \[\]
  
[] are used for indexing

```{r}

# Gets the second thing
x[2] 

# Gets the first row
cars[1, ] 

# Gets the first column
cars[, 1] 

# Gets the value in the first row
# in the first column
cars[1, 1] 


# Gets the first 10 rows of the "dist" column
cars[1:10, "dist"]
```  


## $  
$ is a shortcut used for accessing names within a data.frame

```{r}
cars$speed
```

It is equivalent to:

```{r}
cars[, "speed"]
```


## "" and ''
  
"" and '' are both used to specify strings

```{r}
z = c("string1", 'string2', "string3")
```

## ``
  
`` are used for making something literal

```{r}
cars$`1 $ bad name` = 1:nrow(cars)

head(cars)
```

## ==
  
== is strictly equivalent

```{r}
cars[cars$speed == 20, ]
```

## !=
  

!= is not equal to

```{r}
cars[cars$speed != 20, ]
```


## |  

| is an or operator

```{r}
cars[cars$speed == 18 | cars$speed == 20, ]
```

## &

& means and

```{r}
cars[cars$speed == 18 & cars$dist == 56, ]
```


## %*%
  
%*% is matrix multiplication

```{r}
x %*% y
```

## %%
  
%% is the modulus

```{r}
10 %% 3
```

## %>% 

%>% is a pipe that passes the output of one function into another function:

```{r}
library(dplyr)

1:10 %>% 
  mean()
```

See the magrittr package for more potential pipes.

# Finding Packages

<a href="https://cran.r-project.org/web/views/">CRAN Task Views</a> will help you find packages that do what you want to do. They are arranged into broad categories that then get broken down.

There is also <a href="https://rseek.org/">rseek</a>


# Package Installation

Install new packages from CRAN with:

```{r, eval = FALSE}
install.packages("tidyverse")
```

If you want to install multiple, you need to use a character vector:

```{r, eval = FALSE}
install.packages(c("tidyverse", "psych", "Hmisc"))
```


If you want to install something from Github:

```{r, eval = FALSE}
devtools::install_github("tidyverse/ggplot2")
```


# Loading Packages

If you want to use a package, you load it into your local environment with:

```{r, eval = FALSE}
library(dplyr)
```


# Data Types

There are a few different types of data within R.

We have numeric variables:

```{r}
nums = 1:10

nums
```

We have characters:

```{r}
chars = c("Poor", "Fair", "Good", "Great")

chars
```

And we have factors:

```{r}
facs = as.factor(chars)

facs
```

We can have just regular factors like above or we can order those factors:

```{r}
orderedFacs = ordered(facs, 
                      levels = c("Poor", "Fair", "Good", "Great"))

orderedFacs
```


# Data Structures

At our most basic, we have a vector. Think of it a just a basic list of numbers:

```{r}
vectorExample = rnorm(10)

vectorExample
```

There is the matrix. A matrix is a rectangle where everything is the same type of data:

```{r}
characterMatrix = matrix(letters, nrow = 5, ncol = 5)

characterMatrix
```

```{r}
numericMatrix = matrix(1:25, nrow = 5, ncol = 5)

numericMatrix
```


If we have data of different types, we are dealing with data frames:

```{r}
dfExample = data.frame(numVar = 1:3, 
                       charVar = letters[1:3], 
                       facVar = factor(c("Poor", "Good", "Great")), 
                       ordVar = ordered(c("Poor", "Good", "Great"), 
                                        levels = c("Poor", "Good", "Great")))

summary(dfExample)
```


We also have lists. Lists can contain any number of any thing within the different list entries:

```{r}
listExample = list(1, rnorm(10), numericMatrix, rnorm(30), dfExample)

listExample
```

# Useful Functions

There are many functions that will be helpful, but here are a few easy ones:

```{r}
sum(1:10, na.rm = TRUE)

mean(1:10, na.rm = TRUE)

sd(1:10, na.rm = TRUE)

paste("tic", "tac", "toe", sep = "-")

state.abb

match(c("January", "February"), month.name)


testWords = c("Bad", "Poor", "Great", "Awesome")

dfExample[dfExample$facVar %in% testWords, ]

dfExample[!(dfExample$facVar %in% testWords), ]

```

