---
title: "Data Wrangling With R"
output: revealjs::revealjs_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Wrangling

Data wrangling is potentially comprised of many different tasks.

At the end of the day, we are trying to create rectangles.

## {data-background="https://assets.mubi.com/images/film/18159/image-w856.jpg?1445924140"}

## {data-background="http://theopenews.com/images/articoli/horse.jpg"}

## {data-background="http://wpmedia.o.canada.com/2014/06/martin-tata9.jpg"}

## Why are you here?

## Embed a web page as a background {data-background-iframe="https://www.youtube.com/watch?v=_UQsqFiyw9k"}

## 

If you have used R before, you can completely ignore me for the next few minutes. 

  - Wait for the special signal to return your attention!

Instead of peaking ahead or browsing the internet, feel free to listen to these tunes:


## Basics: Quickly

Everything in R is an object -- everything. single values, models, and even functions -- all objects.

To create an object, we just assign it a name.

```{r}
x = 1:5

y <- 6:10
```

You can use either "=" or "<-".

- The shortcut for "<-" is Alt + -

## Assignment

The stray puppy:

```{r}
1:5
```

Naming the stray:

```{r}
fido = 1:5
```

Grooming the stray:

```{r}
fido = fido * 2
```



## Applying Functions

Now that we have some objects, x and y, we can apply some functions to each:

```{r}

xMean = mean(x)

ySD = sd(y)

```


## Combinging Function

We can also combined functions!

```{r}

abs(mean(-1:-10))

```

Clear as mud, right?

## Packages

Many functions are grouped into things called packages. 

Packages are usually written to contain functions that do similar things.

You load a package like this:

```{r, eval = FALSE}
library(dplyr)
```


If you don't have a package already, you install it with:

```{r, eval = FALSE}
install.packages("dplyr")
```

## Important Issues

All closures need...closed: (), {}, '', ""

Case matters:

  -  thisName != ThisName != this_name
  
## our Packages

These are the packages that you will need:

```{r, eval = FALSE}
install.packages(c("tidyverse", "rio", 
                   "jsonlite", "psych", 
                   "lavaan" "reshape", 
                   "reshape2", "stringdist"))
```

If more creep up, we will install them when the time is right.