---
title: "The Dark Art of Scraping"
author: "A Telephasic Workshop"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: pygments
    center: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE)
```

## Scraping Can Be Easy

You have already made friends with `rvest` and SelectorGadget.

With `rvest`, getting tables and anything within the html is very easy.

## But...

Sometimes you need to dig into the structure to find what you want.

##

What if we just want the names from this <a href="https://mendoza.nd.edu/mendoza-directory/">page</a>?

## 

SelectorGadget might lead us to this:

```{r}
library(rvest)

read_html("https://mendoza.nd.edu/mendoza-directory/") %>% 
  html_nodes(".a-profile") %>% 
  html_text() %>%
  head(10) %>% 
  knitr::kable() 
```

##

Is that what we want?

##

We will need to get a little bit more into the CSS to extract only the needed information.

## Learning CSS

With your free time, you can learn about CSS selectors:

<a href = "https://flukeout.github.io/">The CSS Diner</a>

<a href = "https://www.w3schools.com/cssref/css_selectors.asp">Selector Reference</a>

## Using CSS

If we just want the names, we just need to add a little more information about the nodes.

```{r}
library(rvest)

read_html("https://mendoza.nd.edu/mendoza-directory/") %>% 
  html_nodes(".a-profile:nth-child(1)") %>% 
  html_text() %>% 
  head(10) %>% 
  knitr::kable() 
```


## The Wheels Always Come Off

You might find that CSS will not always get you what you want.

This is happening with greater frequency.

## When CSS Won't Work

Let's look at the college football playoff ticket options <a href="https://www.cfp-rsvp.com/home">page</a>.

Say we want the listed prices.

## 

This should work:

```{r, eval = FALSE}
read_html("https://cfp-rsvp.com/home") %>% 
  html_nodes(".price-current")
```

## But...

```{r,}
read_html("https://cfp-rsvp.com/home") %>% 
  html_nodes(".price-current")
```


## {data-background="https://media.giphy.com/media/YVPwi7L2izTJS/giphy.gif"}

## {data-background="https://media.giphy.com/media/hmE2rlinFM7fi/giphy.gif"}

## Why Did This Happen?

It should have worked, given that we used the appropriate CSS.

The problem is that the information does not really live on that page.

We need to do some additional digging.

##

We need to check the network and look for a specific file.

We don't even need to scrape anything.

## {data-background="https://media.giphy.com/media/VHHxxFAeLaYzS/giphy.gif"}

## json

Some work needs to be done, but you have the information.

```{r, eval = FALSE}
library(jsonlite)

fromJSON("https://api.dibitnow.com/api/v1/eventPerformers/getEventPerformers?eventId=15")
```

It is worth noting that json tables are becoming more common.

So scraping them can be a challenge, but finding the source is easy.

##

Sometimes, json won't even be able to save you.

## Selenium

When things get tough, you can turn to Selenium.

Selenium provides web browser automation. 

## Selenium In Action

```{r, eval = FALSE}
library(purrr)

library(RSelenium)

rd = rsDriver(browser = c("firefox"))

driver = rd$client

driver$navigate("https://www.cfp-rsvp.com/home")

element = driver$findElement("css", "body")

element$sendKeysToElement(list(key = "page_down"))

elements = driver$findElements(using = "css selector", ".price-current")

ticketPrices = map(elements, ~ .x$getElementText()[[1]])
```

## Just Because We Can

```{r, eval = FALSE}
element = driver$findElements(using = "css selector", "a[href*=team]")

driver$mouseMoveToLocation(webElement = element[[1]])

driver$click()

driver$close()

rd$server$stop()
```


## {data-background="https://media.giphy.com/media/26tPnAAJxXTvpLwJy/giphy.gif"}

## Wrapping Up

Through `rvest` or `RSelenium`, you will probably be able to scrape the target data.

Both will let you sign into a webpage and proceed onward. 

You can also pass other information into a website through the headers.

##

In any scraping exercise, proceed with care.

Pay attention to the ToS and don't hit the site too hard.