---
title: "Sentiment Analysis"
output:
  revealjs::revealjs_presentation:
    theme: solarized
    center: true
    transition: fade
    slide_level: 2
---

## Sentiment 

Used when we want to know the general *feelings* of text. 

Applied to Twitter and other social media posts

Can use it anywhere where people have written/said something.

Sentiment can take many different forms: positive/negative affect, emotional states, and even financial contexts.

## 

We will cover two forms of sentiment:

- Word level (simple)

- Sentence level (complex)

## Skipping Around

We are not going to get into text prep now. 

It is practically its own lecture.


## Helpful Packages

```{r, eval = FALSE}
install.packages(c("tidytext", "sentimentr"))
```


## Simple

Let's consider the following statements:

```{r, warning = FALSE, message = FALSE}
library(dplyr); library(tidyr); library(tidytext)

statement <- "I dislike beer, but I really love the shine."

tokens <- tibble(text = statement) %>% 
  unnest_tokens(tbl = ., output = word, input = text)

tokens
```


## 

Using our tokens against a pre-defined dictionary:

```{r, warning = FALSE, message = FALSE}
tokens %>%
  inner_join(get_sentiments("bing")) %>% 
  count(sentiment) %>% 
  spread(sentiment, n, fill = 0) %>% 
  mutate(sentiment = positive - negative)
```


## Thinking About The Output

Do you think that disklike and love are of the same magnitude? 

I might say that love is stronger than dislike. 

Let's switch out our sentiment library to get something with a little better notion of polarity.

##

```{r, warning = FALSE, message = FALSE}
tokens %>%
  inner_join(get_sentiments("afinn"))
```

Now this looks a bit more interesting! "Love" has a stronger positive polarity than "dislike" has negative polarity. So, we could guess that we would have some positive sentiment.

##

We can get an idea of our sentence's overall sentiment, if we divide the sum of our word sentiments by the number of words within the dictionary

```{r, warning = FALSE, message = FALSE}
tokens %>%
  inner_join(get_sentiments("afinn")) %>% 
  summarize(n = nrow(.), sentSum = sum(score)) %>% 
  mutate(sentiment = sentSum / n)
```

##

These simple sentiment analyses provide some decent measures to the sentiment of our text. 

We are ignoring big chunks of our text by just counting keywords.

## Smarter Sentiment Analysis

```{r, warning = FALSE, message = FALSE}
library(sentimentr); library(lexicon); library(magrittr)

statement <- "I dislike beer, but I really love the shine."

sentiment(statement, polarity_dt = lexicon::hash_sentiment_jockers)
```

##

The first part of our sentence starts out negative (dislike has a sentiment value of -1).

We have an adversarial "but" that will downweight whatever is in the initial phrase.

We have the amplified sentiment of "love" (with a weight of .75 in our dictionary).

With all of this together, we get a much better idea about the sentiment of our text.

## An Added Bonus

I want to show you gganimate!

```{r, eval = FALSE}
library(gganimate); library(ggplot2)

ggplot(mtcars, aes(mpg, wt, color = as.factor(cyl))) + 
  geom_point() +
  geom_smooth(method = "lm") +
  scale_color_brewer(type = "qual") +
  theme_minimal() +
  transition_states(cyl)
```


##

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(gganimate); library(ggplot2)

ggplot(mtcars, aes(mpg, wt, color = as.factor(cyl))) + 
  geom_point() +
  geom_smooth(method = "lm") +
  scale_color_brewer(type = "qual") +
  theme_minimal() +
  transition_states(cyl)
```