---
title: "Process Simulation"
description: |
  Examples using simmer
author:
  - name: Seth Berry
    affiliation: QDM
output: radix::radix_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Queueing Theory

Exactly what it sounds like: it is mathematical study of lines. On its face, maybe not super exciting; however, we should consider all of the different things that tend to wait in lines.


## Notation

Queues have their own notation (Kendall's notation), below are some common ones:

$M/D/k$

$M/M/k$

In all of these, the first two letters refer to the mechanism of action.

The first *M* stands for Markov (or memoryless) and denotes that arrivals occur according to a Poisson.

The second *M* in $M/M/k$ means that the job requirement is exponentially distributed.

*D* is deterministic: all jobs require a fixed amount of time.

*k* is the number of servers/workers/etc.

Both of these are generally assumed to have an infinite buffer.

$M/M/1/K$

Here the *K* is specifying a buffer size.

If a queue is $M/D/k$, we can easily compute some helpful statistics (we don't need any fancy software to help us).

$\lambda$ = arrival rate

$\mu$ = service rate

$\rho = \frac{\lambda}{\mu} $ = utilization

Average number of entities in the system is:

$$ L = \rho + \frac{1}{2}\Bigg(\frac{\rho^2}{1 - \rho}\Bigg) $$

Average number in queue:

$$ L_Q = \frac{1}{2}\Bigg(\frac{\rho^2}{1 - \rho}\Bigg) $$

Average system waiting time:

$$ \omega = \frac{1}{\mu}+\frac{\rho}{2\mu(1 - \rho)} $$

Average waiting time in queue:

$$ \omega_Q = \frac{\rho}{2\mu(1 - \rho)} $$


### Markov Chains

Much like flipping a log over will yield an assortment of creatures, peeling back the layers of many methods will reveal a Markov Chain. And just like those assorted creates, you don't really know what you are looking at when you see them.

Here is a conceptual example of markov chains! This example is adapted from Richard McElreath's excellent book, *Statistical Rethinking*.

You manage 10 teams, named from 1 to 10. Each team name is proportional to the number of people on the team and each team is on a separate floor in the building.

You need to visit a team everyday, proportional to the number of people on the team (i.e., you would visit team 10 more often than team 1).

At the end of the day, you randomly select whether a proposed move will take you up a floor or down a floor.

After randomly selecting up or down, you grab a number of pens that is equal to the number of the current team (for team 10, you would grab 10 pens).

Next, you would grab a number of pencils corresponding to the proposed move (your randomly selected proposal would take you up a floor, so starting back at the bottom with team 1). So you would have 10 pens and 1 pencil.

If you have more pencils than pens, you will always move to the proposed floor.

If you have more pens than pencils, you set down a number of pens equal to the number of pencils and put them in a drawer.

You reach back into the drawer to randomly select a pen or a pencil.

Your selection decides where you go -- pen is stay and pencil is go!

```{r}
markovSim = function(daysRun, startDay) {
  
  position = rep(0, daysRun)
  
  current = startDay
  
  for(i in 1:daysRun) {
    position[i] = current
    
    proposal = current + sample(c(-1, 1), size = 1)
    
    if(proposal < 1) proposal = 10
    
    if(proposal > 10) proposal = 1
    
    probabilityMove = proposal / current
    
    current = ifelse(runif(1) < probabilityMove, proposal, current)
    
    # print(paste(position[i], proposal, probabilityMove, current, sep = " -> "))
  }
  
  return(position)
}

test1 = markovSim(1000, 5)

test2 = markovSim(1000, 6)

library(ggplot2)

ggplot() + 
  geom_line(data = as.data.frame(test1), aes(1:length(test1), test1), 
            color = "#ff5500", size = .75) +
  theme_minimal()
```




# Packages 

```{r, eval = FALSE}
install.packages(c("tidyverse", "simmer", "simmer.plot"))
```

```{r}
library(dplyr)

library(simmer)

library(simmer.plot)
```

## The Bank

### Single Counter

```{r}
set.seed(1001)

customer = trajectory("Customer's path") %>%
  log_("Here I am") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("counter") %>%
  log_(function() {paste("Waited: ", now(bank) - get_attribute(bank, "start_time"))}) %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter") %>%
  log_(function() {paste("Finished: ", now(bank))})

bank = simmer("bank") %>%
  add_resource("counter") %>%
  add_generator("Customer", customer, function() {c(0, rexp(4, 1/10), -1)})


bank %>% run(until = 400)

customerInfo = bank %>%
  get_mon_arrivals %>%
  dplyr::mutate(waiting_time = end_time - start_time - activity_time)
```

### Two Counters -- One Line

```{r}
set.seed(1001)

customer = trajectory("Customer's path") %>%
  log_("Enter") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("counter") %>%
  log_(function() {paste("Waited: ", now(bank) - get_attribute(bank, "start_time"))}) %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter") %>%
  log_(function() {paste("Finished: ", now(bank))})

bank = simmer("bank") %>%
  add_resource("counter", 2) %>%
  add_generator("Customer", customer, function() {c(0, rexp(49, 1/10), -1)})

bank %>% run(until = 400)

customerInfo2Counters = bank %>%
  get_mon_arrivals %>%
  dplyr::mutate(waiting_time = end_time - start_time - activity_time)

```



### Two Counters -- Two Lines

```{r}
set.seed(1001)

customer = trajectory("Customer's path") %>%
  log_("Enter") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  select(c("counter1", "counter2"), policy = "shortest-queue") %>%
  seize_selected() %>%
  log_(function() {paste("Waited: ", now(bank) - get_attribute(bank, "start_time"))}) %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release_selected() %>%
  log_(function() {paste("Finished: ", now(bank))})

bank = simmer("bank") %>%
  add_resource("counter1", 1) %>%
  add_resource("counter2", 1) %>%
  add_generator("Customer", customer, function() {c(0, rexp(49, 1/10), -1)})

bank %>% run(until = 400)

customerInfo2Counters = bank %>%
  get_mon_arrivals %>%
  dplyr::mutate(waiting_time = end_time - start_time - activity_time)

```

### Sensible Queue Sizes

```{r}
set.seed(1002)

customer = trajectory("Customer's path") %>%
  log_("Enter") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("counter1", continue = FALSE, reject = trajectory("Balked customer") %>%
          log_("BALKING")) %>% 
  log_(function() {paste("Waited: ", now(bank) - get_attribute(bank, "start_time"))}) %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter1") %>% 
  log_(function() {paste("Finished: ", now(bank))})

bank = simmer("bank") %>%
  add_resource("counter1", capacity = 1, queue_size = 4) %>%
  add_generator("Customer", customer, function() {c(0, rexp(49, 1/10), -1)})

bank %>% run(until = 400)

customerInfo2Counters = bank %>%
  get_mon_arrivals %>%
  dplyr::mutate(waiting_time = end_time - start_time - activity_time)

nBalked = sum(!get_mon_arrivals(bank)$finished)
```

### Multiple Runs

```{r}
library(parallel)

customer = trajectory("Customer's path") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("counter1", continue = FALSE, reject = trajectory("Balked customer")) %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter1")

mclapply(c(1001, 1002, 1003, 1004), function(the_seed) {
  set.seed(the_seed)

  bank = simmer("bank") %>%
    add_resource("counter1", capacity = 1, queue_size = 4) %>%
    add_generator("Customer", customer, function() {c(0, rexp(49, 1/10), -1)})

  bank %>% run(until = 400)
  
result = bank %>%
    get_mon_arrivals %>%
    dplyr::mutate(waiting_time = end_time - start_time - activity_time)
  
paste("Average wait for ", sum(result$finished), " completions was ",
      mean(result$waiting_time), "minutes.")
}) %>% 
  unlist()

```

