---
title: "From Static, To Interactive, To Reactive"
output:
  revealjs::revealjs_presentation:
    theme: solarized
    center: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Packages

You will probably need the following packages: 

```{r, eval = FALSE}
install.packages(c("shiny", "rsconnet", "plotly", "ggplot2"))
```


## Static Plots

Visualizations are just there to see. 

While they can convey a lot of information, they offer no capacity to dig deeper. 

## 

```{r, echo = FALSE}
library(ggplot2)

ggplot(dplyr::starwars, aes(height, mass, color = species)) + 
  geom_point(size = 3.5, alpha = .5) +
  theme_minimal()
```


## Interactive Plots

Interactive plots are useful for letting people play with some portion of the visualization. 

Zooming and hovering are two example of interactions. 

## Plotly Example

```{r, echo = FALSE}
library(plotly)

plot_ly(dplyr::starwars, x = ~height, y = ~mass, color = ~species) %>% 
  add_markers()
```


## Reactive Plots

We can take our interactive plots and add some neat features to them. 

Reactivity responds directly to a users input and changes accordingly. 

##

```{r, echo=FALSE}

plotData <- dplyr::starwars %>% 
  select(height, mass, hair_color, eye_color, birth_year, 
         homeworld, species)

pageWithSidebar(
  headerPanel('Simple Example'),
  sidebarPanel(
    selectInput('xcol', 'X Variable', names(plotData), 
                selected = names(plotData)[[1]]),
    selectInput('ycol', 'Y Variable', names(plotData), 
                selected = names(plotData)[[2]]),
    selectInput('colorCol', 'Color variable', names(plotData), 
                selected = names(plotData)[[7]])
  ),
  mainPanel(
    plotlyOutput('plot1')
  )
)
```

```{r, context="server"}
output$plot1 <- renderPlotly({
    plot_ly(plotData, x = ~get(input$xcol), y = ~get(input$ycol), color = ~get(input$colorCol)) %>% add_markers()
  })
```

## Shiny

Reactive plots are created through `Shiny`. 

It is incredibly flexible and most of what you need can be wrapped into 2 files. 

- server 

- ui 

## server

The server is where all of the wild stuff happens.

## Code Example

```{r, eval = FALSE}
output$plot1 <- renderPlotly({
    plot_ly(plotData, x = ~get(input$xcol), y = ~get(input$ycol), color = ~get(input$colorCol)) %>% add_markers()
  })
```

## ui

The ui does exactly what you might think -- controls the look and the inputs.

## Example Code

```{r, eval = FALSE}
plotData <- dplyr::starwars %>% 
  select(height, mass, hair_color, eye_color, birth_year, 
         homeworld, species)

pageWithSidebar(
  headerPanel('Simple Example'),
  sidebarPanel(
    selectInput('xcol', 'X Variable', names(plotData), 
                selected = names(plotData)[[1]]),
    selectInput('ycol', 'Y Variable', names(plotData), 
                selected = names(plotData)[[2]]),
    selectInput('colorCol', 'Color variable', names(plotData), 
                selected = names(plotData)[[7]])),
  mainPanel(plotlyOutput('plot1')
  )
)
```

## Deployment

You can deploy a Shiny app: 

Locally 

Through a package 

shinyapps.io 

## Demonstrations

The only way to get into Shiny is to play with it -- a lot!