---
title: "Data Visualization"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: zenburn
    center: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(ggplot2); library(lazerhawk); library(plotly); library(dygraphs)
library(dplyr); library(gganimate); library(ggthemes); library(visibly)
library(ggdark)

# The lazerhawk and visibly packages are from github:
# devtools::install_github('m-clark/lazerhawk')
# devtools::install_github('m-clark/visibly')

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  plots <- c(list(...), plotlist)
  numPlots = length(plots)
  if (is.null(layout)) {
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  if (numPlots==1) {
    print(plots[[1]])
  } else {
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    for (i in 1:numPlots) {
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```

## Exploration

Testing hypotheses is a critical part of the scientific process.

But...you need to understand your data first.

- What are some relationships, how are variables distributed, etc.

Visualizing your data can go a long way to helping you explore your data.

## Preattentive Processing

We pick up on visual cues much quicker than reading (i.e., what catches our eye).

- That is why entire disciplines are devoted to visual processing.

Some things hit us pretty quickly:

color (hue & intensity); shape; space/density; line width & orientation; size

## Preattentive Processing In Action

```{r, fig.height = 5.5, fig.width = 10.5, fig.align="center"}
library(dplyr)
library(patchwork)

preProDat <- data.frame(x = rep(1:5, each = 10), y = rep(5:1, times = 10))

preProDat<- subset(preProDat, x != 2 | y != 2)

singlePoint<- data.frame(x = rep(2, 10), y = rep(2, 10))

singleColor <- ggplot(preProDat, aes(x, y)) + 
  geom_point(size = 4) +
  geom_point(data = singlePoint, aes(x, y), color = "red", size = 4) +
  dark_theme_minimal() +
  ggtitle("Color")

multiColor <- ggplot(preProDat, aes(x, y)) + 
  geom_point(size = 4, color = "#ff884d") +
  geom_point(data = singlePoint, aes(x, y), color = "#b33b00", size = 4) +
dark_theme_minimal() +
  ggtitle("Saturation")

singleShape<- ggplot(preProDat, aes(x, y)) + 
  geom_point(size = 4) +
  geom_point(data = singlePoint, aes(x, y), shape = 15, size = 4) +
  dark_theme_minimal() +
  ggtitle("Shape")

singleSize <- ggplot(preProDat, aes(x, y)) + 
  geom_point(size = 4) +
  geom_point(data = singlePoint, aes(x, y), size = 6) +
  dark_theme_minimal() +
  ggtitle("Size")

singleDensity <- ggplot(preProDat, aes(x, y)) + 
  geom_point(size = 4) +
  geom_jitter(data = singlePoint, aes(x, y), size = 4, width = .25, height = .25) +
dark_theme_minimal() +
  ggtitle("Density")

singleOrientation <- ggplot(preProDat, aes(x, y)) + 
  geom_point(size = 4, shape = 124) +
  geom_point(data = singlePoint, aes(x, y), size = 4, shape = 92) +
  dark_theme_minimal() +
  ggtitle("Orientation")

(singleColor | multiColor | singleShape) /
  (singleSize | singleDensity | singleOrientation)
```



## Correlation Matrix

```{r}
library(DT); library(broom)

carCor <- cor(mtcars[, 1:8])

knitr::kable(round(carCor, 2))
```


## Corrplot

```{r, fig.height=5.5, fig.width=9, fig.align="center"}
corrplot::corrplot.mixed(carCor, lower = "number", upper = "square", 
                         order = "hclust")
```

## Heat Map

```{r, fig.height=5.5, fig.width=9, fig.align="center"}
carCor <- as.data.frame(carCor) %>% 
  mutate(item = row.names(carCor)) %>% 
  tidyr::gather(key = key, value = corrVal, -item)

hotCor <- ggplot(carCor, aes(item, key, fill = corrVal)) +
  geom_raster() +
  scale_y_discrete(limits = rev(levels(as.factor(carCor$key)))) +
  scale_fill_gradient2(low = "#3d52a1", mid = "#ffffff", high = "#ae1c3e") +
  visibly::theme_trueMinimal()

ggplotly(hotCor)
```


## More Heat

```{r}
library(GGally)

df <- mtcars[, c(1,3,4,5,6,7)]

ggcorr(df, palette = "RdBu", label = TRUE) +
  dark_theme_minimal()
```

## Inference

Currently, most journals and clients "need" to see test statistics to know if there is a relationship.

- The golden *p*-value is the driver of science!

Not everything needs a test statistic to tell you that a real relationship exists (or not).

## Interocular Traumatic Impact

```{r, fig.width = 10, fig.align="center"}
formula1 <- y ~ x

lmPlot <- ggplot(mpg, aes(cty, hwy)) +
  geom_point(alpha = .75) +
  geom_smooth(method = "lm", formula = formula1, se = TRUE, color = "#44aa99") +
  # ggpmisc::stat_poly_eq(aes(label = ..adj.rr.label..), formula = formula1, 
  #                       parse = TRUE, hjust = -3) +
  theme_bw()

lmPlot2 <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(alpha = .75) +
  geom_smooth(method = "lm", se = TRUE, color = "#44aa99") +
  #geom_smooth(se = FALSE, color = "blue", span = .5) +
  #geom_smooth(method = "gam", formula = y ~ splines::bs(x, 4), se = FALSE, color = "#cc6677") +
  # ggpmisc::stat_poly_eq(aes(label = ..adj.rr.label..), formula = formula1, 
  #                       parse = TRUE, hjust = -3) +
  theme_bw()

multiplot(lmPlot, lmPlot2, cols = 2)

```



## Maybe A Better Fit

```{r, fig.align = "center", fig.height = 5.5, fig.width = 10}
formula1 <- y ~ x
formula2 <- y ~ splines::bs(x, 4)

gamFitPlot <- ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(method = "lm", formula = formula1, se = FALSE, color = "#44aa99") +
  #geom_smooth(se = FALSE, color = "blue", span = .5) +
  geom_smooth(method = "gam", formula = formula2, se = FALSE, color = "#cc6677") +
  # ggpmisc::stat_poly_eq(aes(label = ..adj.rr.label..), formula = formula2, 
  #                       parse = TRUE, color = "#cc6677", hjust = -5) +
  # ggpmisc::stat_poly_eq(aes(label = ..adj.rr.label..), formula = formula1, 
  #                       parse = TRUE, color = "#44aa99", label.y.npc = .9, hjust = -5) +
  theme_bw()

ggplotly(gamFitPlot)
```



## Principles And Properties

## What Was And What Should Never Be

```{r, fig.height=5.5, fig.width=10}
excelExample <- ggplot(mtcars, aes(factor(cyl), fill = factor(am))) +
  geom_bar(position = "dodge") +
  theme_excel() + 
  scale_fill_excel() +
  ggtitle("Excel 2003")

pieExample <- ggplot(mtcars, aes(x = factor(1), fill = factor(cyl))) +
  geom_bar(width = 1) +
  theme_void() +
  scale_fill_excel() +
  coord_polar(theta = "y") +
  ggtitle("This Ain't No Data\nThis Ain't No Disco")

tenRing <- ggplot(mtcars, aes(x = factor(1), fill = factor(cyl))) +
  geom_bar(width = 1) +
  theme_void() +
  scale_fill_excel() +
  coord_polar() +
  ggtitle("The 10-ring Is Too Big")


df <- data.frame(
  variable = c("Nom", "nom"),
  value = c(20, 80)
)

pacMan <- ggplot(df, aes(x = "", y = value, fill = variable)) +
  geom_bar(width = 1, stat = "identity") +
  scale_fill_manual(values = c("yellow", "black")) +
  coord_polar("y", start = pi / 3) +
  theme_void() +
  labs(title = "Wickham's\nPac-Man")

multiplot(excelExample, pacMan, tenRing, pieExample, cols = 2)

```



## Color

Color is generally used for the following:

- Groups

- Intensity


Only use color when it is actually doing something to aid interpretation.


```{r, fig.width = 10, fig.height= 3, fig.align="center"}
no <- ggplot(mpg, aes(class)) + 
  geom_bar(aes(fill = class)) +
 dark_theme_minimal() +
  ggtitle("No") +
  guides(fill = FALSE) + 
  theme(axis.text.x = element_text(angle = 90, vjust = .5))

better <- ggplot(mpg, aes(class)) + 
  geom_bar(fill = "#9ECAE1") +
  theme_trueMinimal() +
  ggtitle("Better") +
  theme(axis.text.x = element_text(angle = 90, vjust = .5))

multiplot(no, better, cols = 2)

```



## Color - Groups

```{r, fig.height = 6, fig.width = 10, fig.align="center"}
ggplot(mpg, aes(factor(class), fill = factor(cyl))) +
  geom_bar(position = "dodge") +
  dark_theme_minimal() +
  scale_fill_ptol()
```


## Color - Intensity

```{r, fig.align="center"}

ggplot(faithfuld, aes(waiting, eruptions, fill = density)) + 
  geom_raster(interpolate = TRUE) + 
  scale_fill_gradient(low = "#f2f0f7", high = "#4a1486") + 
  visibly::theme_trueMinimal()

```


## Color - Brewer Palettes

```{r, fig.height=6, fig.width=10, fig.align="center"}
RColorBrewer::display.brewer.all()
```


## Color - Tol Palettes

```{r, fig.height=6, fig.width=10, fig.align="center"}
scales::show_col(ptol_pal() (6))
```


## Color - Few Palettes



```{r, fig.height=5, fig.width=5, echo = FALSE, eval = FALSE}
scales::show_col(few_pal("Light")(6))
```

```{r, fig.height=5, fig.width=5}
scales::show_col(few_pal("Dark")(6))
```



## Size

Much like color, size can convey very distinct ideas:

- Value size

- Frequency

You need to be very explicit to which you are referring!


## Size


```{r, fig.height=5, fig.width=10, fig.align="center"}
valSize <- ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(size = qsec), color = "#117777", alpha = .5) + 
  dark_theme_minimal() 

countSize <- ggplot(mpg, aes(class, as.factor(cyl))) +
  geom_count(color = "#117777") +
  scale_size_area(breaks = c(1, 10, 20, 30)) +
  dark_theme_minimal() 

multiplot(valSize, countSize, cols = 2)

```



## Shape

Shape is a bit more limited than color and size.

- It really only helps to denote group membership.

  - And only does well when things are sparse.

## Shape - Linetypes {.flexbox .vcenter}

```{r}
df <- data.frame(
  x = rep(1:10, 4), 
  y = 1:40,
  group = rep(c("a", "b", "c", "d"),
  each = 10)
)

ggplot(sample(df, 100, replace = TRUE), aes(x=x, y=y, group=group)) +
  geom_line(aes(linetype = group), color = "#781c81", size = 1.25) +
  theme_bw()
```

## Where We Have Trouble...


```{r, fig.height=5, fig.width=9, fig.align="center"}
df <- data.frame(
  x = rep(1:10, 4), 
  y = sin(1:40),
  group = rep(c("a", "b", "c", "d"),
  each = 10)
)

badNews <- ggplot(sample(df, 100, replace = TRUE), aes(x=x, y=y, group=group)) +
  geom_line(aes(linetype = group), color = "#781c81", size = 1.25) +
  theme_bw() +
  ggtitle("Um....What?")

slightlyBetter <- ggplot(sample(df, 100, replace = TRUE), aes(x=x, y=y, color = group)) +
  geom_line(size = 1.25, alpha = .75) +
  dark_theme_minimal() +
  scale_color_ptol() +
  ggtitle("Better...Marginally")

multiplot(badNews, slightlyBetter, cols = 2)
```


## Shape - Points

```{r, fig.align="center"}
invert_geom_defaults()
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape = factor(cyl)), size = 4) +
  theme_bw()
```


## Alpha

> - No, we are not talking about Cronbach.

> - We probably all know about the RGB color model.

> - But...did you know that it can have a fourth channel?

> - Alpha!

> - Alpha is essentially transparency.

> - 0 is completely translucent

## How Does Alpha Help

We are frequently fortunate enough to have a considerable amount of data.

Depending on what type of plotting we want to do, we could run into problems very quickly.


## Alpha


```{r, fig.height=5.5, fig.width=10, fig.align="center"}
alpha0 <- ggplot(diamonds, aes(carat, price)) +
  geom_point() +
  theme_bw() +
  ggtitle("Alpha = 1")
  
alpha5 <- ggplot(diamonds, aes(carat, price)) +
  geom_point(alpha = .5) +
  theme_bw() +
  ggtitle("Alpha = .5")

alpha1 <- ggplot(diamonds, aes(carat, price)) +
  geom_point(alpha = .1) +
  theme_bw() +
  ggtitle("Alpha = .1")

alpha01 <- ggplot(diamonds, aes(carat, price)) +
  geom_point(alpha = .01) +
  theme_bw() +
  ggtitle("Alpha = .01")

multiplot(alpha0, alpha5, alpha1, alpha01, cols = 4)  
```


## Jitter

If you have ever held a jumping bean or drank too much coffee, you might already know about jitter.

Jitter is useful if we have some minor overplotting issues.

- It would not be a good fix for what we just saw.

```{r, fig.height=3.8, fig.width=10}
ggplot(diamonds, aes(carat, price)) +
  geom_jitter() +
  theme_bw() +
  ggtitle("I Told You It Wouldn't Work")
```


## Looks Pretty Clear... {.flexbox .vcenter}

```{r, fig.height=5.5, fig.width=9}
ggplot(mpg, aes(as.factor(cyl), hwy)) + 
  geom_point() + 
  theme_bw()
```


## After Jittering {.flexbox .vcenter}
  
```{r, fig.height=5.5, fig.width=9}
ggplot(mpg, aes(as.factor(cyl), hwy)) + 
  geom_jitter(width = 0.4, height = 0) + 
  theme_bw()
```


## An Alternative {.flexbox .vcenter}

```{r, fig.height=5.5, fig.width=9}
ggplot(mpg, aes(as.factor(cyl), hwy)) + 
  geom_boxplot(fill = "#ff5500") + 
  theme_bw() 
```


## Other Ideas

```{r, fig.height=5.5, fig.width=10, fig.align="center"}

sizedPlot <- ggplot(mpg, aes(as.factor(cyl), hwy)) + 
  geom_count(alpha = .75) + 
  theme_bw() +
  ggtitle("Size")

densityEst <- ggplot(mpg, aes(hwy, fill = as.factor(cyl))) + 
  geom_density(alpha = .5) + 
  theme_bw() +
  scale_color_ptol() +
  ggtitle("Density Estimation")

multiplot(sizedPlot, densityEst, cols = 2)
```


## Small Multiples/Faceting

We often want to compare a plot over several different levels. 

We have already seen ways to use color and other items to make groups distinct.

If you want to be daring, though, there are other methods.

## Small Multiples/Faceting {.flexbox .vcenter}

```{r, fig.height=5.5, fig.width=9}
ggplot(mpg, aes(displ, hwy)) + geom_point() +
  facet_grid(. ~ cyl) + theme_bw()
```


## Multiple Relationships -- Very Powerful {.flexbox .vcenter}

```{r, fig.height=5.5, fig.width=10}
ggplot(mpg, aes(displ, cty)) + 
  geom_count(alpha = .3) +
  geom_rug(position = "jitter") +
  facet_grid(drv ~ cyl) + 
  theme_bw()
```


## Combining Features

We just saw a glimpse about how we can use multiple principles to convey information.

 - We had a two-variable facet, size, and density estimates.

```{r, fig.height=3, fig.width=4.75}
ggplot(diamonds, aes(carat, price)) +
  geom_point(alpha = .1) +
  geom_rug() +
  visibly::theme_trueMinimal() +
  ggtitle("Uninformative Margins")

```


##

```{r, fig.height=3, fig.width=4.75}
p <- ggplot(diamonds, aes(carat, price)) +
  geom_point(alpha = .1) +
  theme_trueMinimal() +
  ggtitle("Informative Margins")

ggExtra::ggMarginal(p)
```


## Combining Features {.flexbox .vcenter}

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(size = qsec, color = as.factor(cyl), shape = as.factor(gear))) + 
  geom_rug(position = "jitter") +
  theme_bw() + 
  geom_smooth(se = FALSE) +
  facet_grid(. ~ am)
```


## Other Combinations

```{r, fig.align="center"}
p <- ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() + 
  geom_rug(position = "jitter") +
  theme_bw()

ggExtra::ggMarginal(p, type = "boxplot")
```


## Tips (And Maybe Tricks)

## Grid Lines

The default is always some version of "No".

```{r, fig.width = 10, fig.align = "center"}
full <- ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(size = 4) +
  theme(panel.grid.major = element_line(color = "black", size = 1), 
        panel.grid.minor = element_line(color = "black", size = 1), 
        panel.background = element_rect(fill = "white")) +
  ggtitle("Is It A Map?\nMaybe Graphing Paper?")

slim <- ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(size = 4) +
  theme_bw() +
  ggtitle("Better")

none <- ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(size = 4) +
  visibly::theme_trueMinimal() +
  ggtitle("Data!")

multiplot(full, slim, none, cols = 3)

```


## When To Use Grid Lines

Use for:

- Large plots
- Make differences clear
- Comparing values on categorical scales
- Narrowing focus

When you do use them, they should be very faint.


## Large Plots

```{r, fig.width = 10, fig.height = 5.75}
noLines <- ggplot(economics, aes(date, unemploy)) +
  geom_line() +
  visibly::theme_trueMinimal() +
  ggtitle("Maybe Tough To Follow")

lightLines <- ggplot(economics, aes(date, unemploy)) +
  geom_line() +
  theme_light() +
  ggtitle("Pretty Clear")

multiplot(noLines, lightLines)

```


## Make Differences Clearer

```{r, fig.width = 10, fig.height = 5.75}
noLines <- ggplot(mpg, aes(class, hwy)) +
  geom_boxplot() +
  visibly::theme_trueMinimal() +
  coord_flip() +
  ggtitle("Hard To Judge Differences")


lightLines <- ggplot(mpg, aes(class, hwy)) +
  geom_boxplot() +
  theme_light() +
  coord_flip() +
  ggtitle("Easier")

multiplot(noLines, lightLines, cols = 2)

```


## Comparing Categorical Scale

```{r, fig.width = 10, fig.height = 5.5}
df <- data.frame(
  x = c(cos(1:10), sin(rep(1:10, 2)), cos(1:10)), 
  y = 1:40,
  group = rep(c("a", "b", "c", "d"),
  each = 10)
)

noLines <- ggplot(sample(df, 100, replace = TRUE), aes(x=x, y=y, group=group)) +
  geom_line(aes(color = group), size = 1.25) +
  visibly::theme_trueMinimal() +
  scale_color_ptol() +
  ggtitle("Hard To Compare Groups")

lightLines <- ggplot(sample(df, 100, replace = TRUE), aes(x=x, y=y, group=group)) +
  geom_line(aes(color = group), size = 1.25) +
  theme_light() +
  scale_color_ptol() +
  ggtitle("Infinitely Easier")

multiplot(noLines, lightLines)
```



## Narrowing Focus

```{r, fig.width = 10, fig.height = 6}
noLines <- ggplot(mpg, aes(hwy, cty)) +
  geom_point(alpha = .5) +
  visibly::theme_trueMinimal() +
  ggtitle("Finding Chunks Is Tricky")

lightLines <- ggplot(mpg, aes(hwy, cty)) +
  geom_point(alpha = .5) +
  theme_light() +
  ggtitle("Pretty Easy Here")

multiplot(noLines, lightLines)
```



## The Correct Tool For The Job

## Continious and Categorical

## Jitter

```{r}
ggplot(diamonds, aes(as.factor(cut), price)) +
  geom_jitter() +
  theme_minimal()
```

## Boxplots

```{r}
ggplot(diamonds, aes(as.factor(cut), price)) +
  geom_boxplot() +
  theme_minimal()
```

## Ridges

```{r}
ggplot(diamonds, aes(x = price, y = as.factor(cut), fill = ..x..)) +
  ggridges::geom_density_ridges_gradient() +
  scale_fill_viridis_c() +
  theme_minimal()
```


## Violin Plots

```{r}
ggplot(diamonds, aes(as.factor(cut), price)) +
  geom_violin() +
  theme_minimal()
```

## Sina/Beeswarm Plots

```{r}
library(ggforce)

ggplot(diamonds, aes(as.factor(cut), price)) +
  geom_violin() +
  geom_sina() +
  theme_minimal()
```

## Categorical and Categorical

```{r}
diamonds %>% 
  group_by(cut, color) %>% 
  summarize(count = n()) %>% 
  ggplot(., aes(cut, color, fill = count)) +
  geom_tile() +
  dark_theme_minimal()
```


## An Excellent Resource

<a href="https://serialmentor.com/dataviz/index.html">Claus Wilke's Book</a>

## Important People

Mike Bostok -- D3

Cynthia Brewer -- Color Brewer

William Cleveland -- Pioneer (Citation Champ)

Stephen Few -- All Things Viz

Michael Friendly -- Psychologist and Viz Expert

Paul Tol -- Physicist and Viz Expert

Edward Tufte -- Information Display

Hadley Wickham -- A million things

Leland Wilkinson -- The Grammar of Graphics
