---
title: "Generalized Linear Models"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    css: documentCSS.css
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = FALSE, comment = "")
```


# On Centering

Let's return to our crime data and one of our models:

```{r}
crimeScore = readr::read_csv("http://nd.edu/~sberry5/data/crimeScore.csv")
```


Let's look at the summary of a few variables:

```{r}
library(dplyr)

crimeScore %>% 
  select(SSL_SCORE, WEAPONS_ARR_CNT, NARCOTICS_ARR_CNT) %>% 
  summary()
```

We see a minimum of 1 for both weapons and narcotics, a mean of 1.2 for weapons, and a mean of 2.06 for narcotics.


```{r}
twoVars = lm(SSL_SCORE ~ WEAPONS_ARR_CNT + NARCOTICS_ARR_CNT, data = crimeScore)

summary(twoVars)
```


If we mean center our predictors, we have a meaningful intercept!

```{r}
crimeScore %>% 
  mutate(weaponsCenter = WEAPONS_ARR_CNT - mean(WEAPONS_ARR_CNT, na.rm = TRUE), 
         narcCenter = NARCOTICS_ARR_CNT - mean(NARCOTICS_ARR_CNT, na.rm = TRUE)) %>% 
  lm(SSL_SCORE ~ weaponsCenter + narcCenter, data = .) %>% 
  summary()
  
```

We can now say that at mean values of weapons and narcotics arrest, the SSL_SCORE is 319.

You don't have to center on just means -- you can choose whatever value might be meaningful.


# Generalized Linear Models

What a wonderful circular definition!

- GLM is telling you exactly what it is within the title.

But, how is it generalized?

## Distributions...

Remember how linear models really enjoy the whole Gaussian distribution scene?

- Not all data follows a Gaussian distribution.

So, we need a way to incorporate different distributions of the dv into our model.

Distributions cannot do it alone!

## And Link Functions

From a theoretical perspective, link functions are tricky to get your head around.

- Find the exponential of the response's density function and derive the canonical link function...

From a conceptual perspective, all they are doing is allowing the linear predictor to "link" to a distribution function's mean.

- If you know a distribution's canonical link function, that is all the deeper you will probably every need.


At the end of the day, these link functions will do *something* to convert the outcome (dv) to an unbounded continuous variable.

## Linear Regression

We already saw linear regression.

A linear regression deals with real numbers between $-\infty$ to $\infty$ in the dependent variable.

It is the most vanilla within the GLM.

- Gaussian distribution with an "identity" link function

## In R

```{r}
lmTest = glm(SSL_SCORE ~ WEAPONS_ARR_CNT, data = crimeScore, family = gaussian)

summary(lmTest)
```

We see a lot of the same information, but we have a new model fit statistic to look at: deviance.

Deviance is *badness*. The null deviance is telling us how well our dependent variable is predicted by a model with only the intercept -- this is generally the case with all null models.

Our model, captured in the residual deviance, showed a marked decrease in deviance with very little lost in degrees of freedom.

We can test it with the following:

```{r}
1 - pchisq(lmTest$deviance, lmTest$df.residual)
```


The value of AIC is not really interpretable -- it is used only for model comparison.

We can also produce our confidence intervals:

```{r}
confint(lmTest)
```


### A Quick Aside On CIs

Confidence intervals express that 95% of samples will contain the true value, which we can never really know.

They are not related to the probability of a result being correct.

They are not a 95% chance that the true estimate falls within.

They are not meant to say that the mean will fall into the confidence interval 95% of the time.

"Doesn't contain 0" does not apply to everything.

## Logistic Regression

Logistic regression is substantially different than linear regression.

- It is also a bit confusing, because it is named after its link function.

Instead of that nice continuous dv, we are dealing with a binomially distributed dv.

- 0's and 1's as responses.
- no $\mu$ or $\sigma^2$; instead we have *p* and *n* 


```{r}
kickstarter = readr::read_csv("https://www.nd.edu/~sberry5/data/kickstarter.csv")

kickstarter = kickstarter %>% 
  filter(state == "successful" | state == "failed") %>% 
  mutate(state = ifelse(state == "successful", 1, 0))
```


```{r}
library(ggplot2)

ggplot(kickstarter, aes(state)) +
  geom_bar() +
  theme_minimal()
```

And a table:

```{r}
addmargins(table(kickstarter$state))
```

If we take the number of successes and divide by the total, we will get the probability of being funded:

```{r}
113081 / 281302
```



## Continuous

```{r}
logTest = glm(state ~ backers, data = kickstarter, 
              family = binomial)
```

We are now dealing with log odds in the coefficients.

- For every unit increase in backers, the log odds of a campaign being funded is  ~.03.

```{r}
summary(logTest)
```

# Some Logistic Grounding

## Log Odds

As its default, R produces the log odds for the regression coefficient in a logistic model. 

The odds portion of this is the probability.  We can look at the relationship between the probability, odds, and log odds.

```{r}
probabilityList = c(.001, .01, .15, .2, 
                    .25, .3, .35, .4, .45, 
                    .5, .55, .6, .65, .7, 
                    .75, .8, .85, .9)
```


We have our list of probabilities (always between 0 and 1).  Now, let's write a function to convert them to odds. We will use the $\\p\, / 1 - p$ equation.

```{r}
oddsConversion = function(p) {
  res = p / (1 - p)
  return(res)
}

odds = oddsConversion(probabilityList)

plot(probabilityList, odds)
```

Now, we can convert them to log odds:

```{r}
plot(odds, log(odds))
```


### The Intercept

The intercept is offering the log odds of a campaign with 0 backers being funded -- converting this back to probability (just adding the exponentiation into solving back to probability), we get the following:

```{r}
exp(coef(logTest)["(Intercept)"]) / (1 + exp(coef(logTest)["(Intercept)"]))
```

We are dealing with a pretty small probability that a campaign with 0 backers could be successful.

### The Predictor

Recall that the coefficient in log odds for the backers was .02645174.

Let's start at the median value of backers:

```{r}
median(kickstarter$backers)
```

Now, let's solve our equation for that one. This will produce the conditional logit.

```{r}
medLogit = logTest$coefficients["(Intercept)"] + 
  (logTest$coefficients["backers"] * median(kickstarter$backers))

names(medLogit) = NULL

medLogit
```

Let's do the same thing for the next sequential value of backers:

```{r}
medLogitPlus1 = logTest$coefficients["(Intercept)"] + 
  (logTest$coefficients["backers"] * (median(kickstarter$backers) + 1))

names(medLogitPlus1) = NULL

medLogitPlus1
```

Now we have two sequential conditional logits that we can subtract from each other:

```{r}
backersCoef = medLogitPlus1 - medLogit

backersCoef
```

This is exactly the coefficient that we got from the model. If we exponentiate the log odds, we are "unlogging" it to get the odds ratio that we saw eariler (just plain old odds at this point):

```{r}
exp(backersCoef)
```

For every unit increase in backer, we have a 2% increase in the odds that the campaign will be successful. These odds are stacking as we increase (see the plot below). When we look at the odds, anything above 1 is in favor of moving from the "0" category to the "1" category, whereas anything below 1 is in favor of not moving from the "0".  From a technical question perspective, this makes absolute sense -- we would expect that having more backers contributes to a better chance of funding.

You can also take those odds back to probability:

```{r}
exp(backersCoef) / (1 + exp(backersCoef))
```



## Putting It Together (Hopefully)

When we do our linear model, we are trying to predict the response -- not the case with logistic regression.  Instead we are trying to predict the probability of going from 0 to 1 (not funded to funded).  So, normal plots that we might usually create with linear models will do us no good here.

First, we need to make our predicitions and put them back into the data:

```{r}
library(dplyr)

kickstarter = kickstarter %>% 
  mutate(predictedProbs = predict(logTest, type = "response"))
```

The predict() function is how we get the predictions from any model.  Do note the type arguement with the "response" specification; this ensures that we are using the type of response particular to the model.

Now, we can plot it:

```{r}

kickstarter %>% 
  filter(backers < 1000) %>% 
  ggplot(., aes(backers, predictedProbs)) +
  geom_line(size = 1.5) +
  theme_minimal()
```

For our data, once you start to get above 250 backers, there is a probablity of 1 that it will be successful.

If we wanted to apply new data to our model, we would also use the predict() function, but give it new data.


## Categorical

Let's pick a few categories to work on.

```{r}
catData = kickstarter %>% 
  select(state, main_category) %>% 
  filter(main_category == "Film & Video" |
           main_category == "Music" |
           main_category == "Games")
```


Let's look a the crosstabs for those variables:

```{r}
addmargins(table(catData))
```


We can start to look at the odds for each category being successful:

```{r}
filmOdds = (21404 / 51057) / (29653 / 51057) # You can reduce this to 21404 / 29653

gamesOdds = (9385 / 22398) / (13013 / 22398)

musicOdds = (21763 / 40956) / (19193 / 40956)
```

And we can take each one back to a probability:

```{r}
filmOdds / (1 + filmOdds)

gamesOdds / (1 + gamesOdds)

musicOdds / (1 + musicOdds)
```

We cam convert our odds to probabilities and back like this:

```{r}
filmOdds / (1 + filmOdds) / (1 - (filmOdds / (1 + filmOdds)))

musicOdds / (1 + musicOdds) / (1 - (musicOdds / (1 + musicOdds)))
```


We could also make statements like this: the odds for a film being successful are about 21 to 29.7, where the odds for music are 21.7 to 19.

We could also compare those two directly:

```{r}
(21763 / 19193) / (21404 / 29653)
```

So the odds that a music campaign will be funded are about 57% higher than the odds for a film.


And we could also compare film to games:

```{r}
(9385 / 13013) / (21404 / 29653)
```

Here, the odds that a game will be funded are less than 1% lower than the odds for a film.


Let's run our logistic regression now:

```{r}
categoryLogTest = glm(state ~ main_category, data = catData, family = binomial)

summary(categoryLogTest)
```


```{r}
exp(coef(categoryLogTest))
```

```{r}
filmOdds
```

The film group is about 28% less likely to be funded.

## Cautions About Logistic Regression

Bivariate relationships are really important for logistic models.

- Empty cells can wreck a model.

Your model needs to see some dispersion of values over the bivariate tables.

- Otherwise, you get what is known as separation (perfect prediction over some levels)

Logistic regression requires a larger sample size than what a linear regression needs.

- The "exact test" is a small sample alternative.

## A Big "Gotcha"

$R^2$ does not apply to a logistic regression.

- There are many pseudo-$R^2$, but they really do not mean the same thing as in linear regression.

- You might be asked for them and many people present them.


# Poisson Regression

Poisson regression is something completely different.

It is used to predict counts.

## The (Sometimes) Thin Line

This gets into an area where we need to think long and hard about our dependent variable and what it actually might be.

- Since Poisson regression gets its name from the Poisson distribution, we should probably see if it follows the Poisson distribution.

## Checking...

```{r}
library(vcd)

shroudData = readr::read_csv("https://www.nd.edu/~sberry5/data/shroudData.csv")

poissonTest = goodfit(shroudData$shroudsProduced, type = "poisson")

summary(poissonTest)
```

This is a $\chi^2$ to test if the distribution deviates from a poisson.

- It probably does not.

# Poisson and Its Friends

For models of this nature (our dependent variable is a count variable), we may have two different distributions with which to operate: the poisson distribution or the negative binomial distribution.

Let’s check this out (it will be important later on!).

```{r}
library(dplyr)

shroudData %>% 
  dplyr::select(shroudsProduced, employeeCount) %>% 
  group_by(employeeCount) %>% 
  summarize(mean = mean(shroudsProduced), var = var(shroudsProduced))
```


What is the purpose of this? We are checking the conditional means and variances. Why is this important? If our variances are larger than our means, we have “over dispersion”. We would expect values to be distributed over levels, but if they are really spread out, this qualifies as over dispersion – this is not good for our poisson model because it will cause downward bias (bias, while not tricky conceptually, presents interesting thought questions).

It looks like everything is mostly okay (for now), so let’s proceed onward with our poisson model:

```{r}
poissonTest = glm(shroudsProduced ~ employeeCount, 
                  data = shroudData, 
                  family = poisson)

summary(poissonTest)
```



```{r}
library(AER)

dispersiontest(poissonTest)
```


**Important Note:** We are going to interpret this almost the same as a linear regression. The slight wrinkle here, though, is that we are looking at the log counts. In other words, an increase in one employee leads to an expected log count increase of ~.029. Just like our logisitc regression, we could exponentiate this to get 1.029189 – every employee we add gets us a 3% increase in shrouds produced. Let’s see what this looks like in action:

```{r}
shroudData = shroudData %>% 
  mutate(predValues = predict(poissonTest, type = "response"))

library(ggplot2)

ggplot(shroudData, aes(employeeCount, predValues)) + 
  geom_count() +
  scale_size_area() +
  theme_minimal()
```


Finally, we can look at the residual deviance (it is comparing our model to a model with perfect prediction) to get at our model fit:

```{r}
pchisq(poissonTest$deviance, poissonTest$df.residual, lower.tail = FALSE)
```


This is a *p*-value – it should not be significant.

With everything coupled together, we have a meaningful coefficient, a clear plot, and adequate model fit. Therefore, we might conclude that there is a positive relationship between number of employees on shift and shrouds produced.


# Zero-inflated (ZIP)

Sometimes we have a seeming abundance of zero values within our data. We can have employees with zero absence periods, lines with zero quality failures, and days without safety issues. What is the process that generated the zeros? Are they coming from our count model (“true” zeroes) or something else (some random process)? This is where zero-inflated models become important. ZIP models are mixture models. We are not going to dive too deeply into this, but all you need to know is that a mixture model contains a “mixture” of different distributions.

```{r}
redlights = readr::read_csv("https://www.nd.edu/~sberry5/data/redlights.csv")

poissonRedlight = glm(citation_count ~ as.factor(camera_installed_ny), 
            data = redlights, 
            family = poisson)

summary(poissonRedlight)
```

With this output, we are comparing citation counts against intersection without a camera and those with a camera.

We see that our coefficient is -.88529 – this means that having a camera leads to having .88529 less log counts than without having a camera.

We can also exponentiate that value to get an indicent rate:

```{r}
exp(poissonRedlight$coefficients["as.factor(camera_installed_ny)1"])
```

Now, we could say that the incident rate (redlight citations) for intersections with a camera is -.42 times less than for intersections without a camera.

If, however, we take a look at citation_count’s distribution, we will see more than a few 0’s.

For our redlight data, we saw that having a camera present had an effect on citations, but would it cause 0 citations? Or might there be something else contributing to the 0’s (e.g., no cars going through that intersection due to construction, no police nearby)? If there are no cars going through the intersection due to construction, is there even a chance of obtaining a non-zero response?

```{r}
library(pscl)

zipTest = zeroinfl(citation_count ~ as.factor(camera_installed_ny),
                   dist = "poisson", data = redlights)

summary(zipTest)
```


# Negative Binomial


Remember that whole issue with our conditional means and standard deviations? If we would have had problems those means and sds, we would need to abandon our poisson distribution in favor of the negative binomial. The poisson distribution works when the sample mean and variance are equal – the negative binomial distribution frees that constraint and allows them to vary freely.

Remember this:

```{r}
redlights %>% 
  dplyr::select(citation_count, camera_installed_ny) %>% 
  group_by(camera_installed_ny) %>% 
  summarize(mean = mean(citation_count), var = var(citation_count))
```

Those look like the start of problems. Let’s check our whole sample now:

```{r}
mean(redlights$citation_count)
```


```{r}
var(redlights$citation_count)
```


```{r}
library(MASS)

nbTest = glm.nb(citation_count ~ as.factor(camera_installed_ny), data = redlights)

summary(nbTest)
```

