---
title: "The General Linear Model"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    css: documentCSS.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## glm...not GLM

The general linear model is one of the foundations of statistics.

The most important thing that you need to know is that it comprises three major techniques:

> - Regression

> - T-tests

> - ANOVA

> - Guess what!

## Assumptions

Linear relationship

- Pretty clear

Multivariate normality

- Remember -- this is not on the raw variables but on the residuals

Little multicollinearity

- Predictors need to be independent from each other

No auto-correlations

- Observations are independent from previous observations

Homoscedasticity

- Equal errors along the regression line

## Assumptions

Assumptions are very important.

They are not, however, statistical death sentences.


## Assumptions According To Gelman

1.  Validity

2.  Linearity

3.  Independence of errors

4.  Equal variance of errors

5.  Normality of errors


## Modern Approaches

> - Has anyone ever heard to transform your variables to make them normal?

> - What about dropping outliers?

> - Neither belongs in modern statistics


## Rules Of Thumb

> - How many people do you need?

> - 20 records per predictor...

> - Not happening here!

> - We will use a more appropriate approach.

## Power Analysis

Do you want to melt most people's brains?

>-  Don't use rules of thumb!

>-  Instead of trusting outdated advice, use actual science to determine how many people you need to find if a difference exists.

## Power Analysis

We need three of the following parameters:

>-  Effect size

>-  Sample size

>-  Significance level

>-  Power

>-  We **should** always be doing this *a priori*

>-  Sometimes, it is fun to be a "statistical coroner"

## Effect Sizes

We won't get too deep here, just know that it is a way to determine the difference between groups.

- It's an improvement of *p*-values alone.

Cohen's *d* is likely the most common.

Let theory be your guide.

> - Be realistic!

## Power

Power is ability to detect an effect.

- In NHST words, we are trying to determine if we correctly reject the null hypothesis.

- Type I errors: Reject a true $H_{o}$ (false positive)

- Type II errors: Reject a false $H_{o}$ (false negative)

>- Which is more dangerous?

## Putting It All Together

Let's use the <span class="func">pwr</span> package.

```{r}
library(pwr)

pwr.f2.test(u = 1, v = NULL, f2 = .05, power = .8)

```

In the function: 

- u is the numerator df (*k* - 1)

- v is the denominator df (*n* - *k*) 

- f2 is signficance level

- \(\Pi = 1 -\beta\)

- \(\beta = Type\,II_{prob}\)


## Your Turn!

Use various values to do an *a priori* power analyses.

How does the proposed sample size change as the number of predictors goes up?

What if you tweak the significance level?

What about power?


## Different Test, Different Power Tests

We just did a test for a linear regression model.

Here is one for a *t*-test:

```{r}
tPower = pwr.t.test(n = NULL, d = 0.1, power = 0.8, 
                    type= "two.sample", alternative = "greater")

plot(tPower)
```


## Congratulations

>-  Your knowledge now far exceeds most professionals!

# Regression

## Do You Know...OLS

Linear regression is likely the singular most important statistic.


## A Demonstration

```{r}
# predictors and response
N = 100 # sample size
k = 2   # number of desired predictors
X = matrix(rnorm(N*k), ncol=k)  
y = -.5 + .2*X[,1] + .1*X[,2] + rnorm(N, sd=.5)  # increasing N will get estimated values closer to these

dfXy = data.frame(X,y)

plot(dfXy$y, dfXy$X1)
```


```{r}
lmfuncLS = function(par, X, y){
  # arguments- par: parameters to be estimated; X: predictor matrix with intercept 
  # column, y: response
  
  # setup
  beta = par                                   # coefficients
  
  # linear predictor
  LP = X%*%beta                                # linear predictor
  mu = LP                                      # identity link
  
  # calculate least squares loss function
  L = crossprod(y-mu)
}
```

```{r}

X = cbind(1, X)

init = c(1, rep(0, ncol(X)))

names(init) = c('sigma2', 'intercept','b1', 'b2')
```


```{r}
optlmLS = optim(par = init[-1], fn = lmfuncLS, 
                X = X, y = y, control = list(reltol = 1e-8))

optlmLS$par
```


```{r}
modlm = lm(y~., dfXy)

summary(modlm)
```


```{r}
# Example
QRX = qr(X)
Q = qr.Q(QRX) # Orthogonal matrix
R = qr.R(QRX) # Upper triangle
Bhat = solve(R) %*% crossprod(Q, y)
qr.coef(QRX, y)
```


```{r}
coefs = solve(t(X)%*%X) %*% t(X)%*%y
```


## Important Terms

There are a few important terms to keep in mind:

>-  Coefficients

>-  Standard Errors

>-  Residuals

## Goodness Of Fit

>- $R^2$

>-  *F*-tests

## $R^2$

Variance accounted for by the predictors.

- Adjusted $R^2$ is just a penalized version.

\(R^2 = 1 -\frac {SS_{res}} {SS_{tot}}\)

Sums of squares play a very important roll in all of this.

>-  Don't worry about know how to compute them.

>-  Just know that they are important


## Let's Just Get On With It

```{r, eval = TRUE}
basicExample = lm(mpg ~ wt + qsec, data = mtcars)
summary(basicExample)
```

## What Does This Mean For Me?

The intercept is the fitted value of mpg when everything else is equal to 0.

The coefficient for "wt" is saying that for every unit increase in pounds, mpg goes down by -5.

- Holding qsec constant

The coefficient for "qsec" is saying that for every unit increase in time, mpg goes up by ~1.

- Holding wt constant

What do these patterns mean?

Do they make sense?

## Let's See How This Holds Up

${mpg} = \alpha + \beta_{1} wt_{t} + \beta_{2}  qsec_{t} + \epsilon$

```{r}
basicExample$coefficients['(Intercept)'] + 
  basicExample$coefficients['wt'] * mtcars$wt[1] +
  basicExample$coefficients['qsec'] * mtcars$qsec[1]


mtcars$mpg[1]
```



# T-tests

## What Are They Good For

You can use a *t*-test to test differences between two groups.

There are two general forms of the *t*-test:

>-  Independent

>-  Paired

## Our Focus

We are going to focus mostly on comparing independent samples.

Unless you are going to be doing experimental work, you will probably not need to use paired tests.

Furthermore, you probably won't ever really need to compare a sample to the population (requires you to know $\mu$)

## Tails

Like many other tests, the *t*-test can be tested with either one tail or two tails.

Alternative hypotheses can be any one of the following:

- $\neq$

- $>$

- $<$


What is the difference?

>-  Do you want to look like you know what you are doing or not?

## One Or Two

In all seriousness, let's consider the following plot:

```{r, eval = TRUE}
hist(rnorm(100000))
```


## Let's Give It A Try

```{r}
t.test(mtcars$mpg ~ mtcars$am, alternative = "two.sided")
```

Try it with different values for alternative and with var.equal = TRUE


## Multiple Tests

When conducting *t*-tests, you will often encounter the need to do a lot of them.

This leads to inflated rates of Type I errors

Many corrections exist:

- Bonferroni

- Tukey

- Scheffé

# Analysis Of Variance

## ANOVA

ANOVA is a lot like a *t*-test, but you can have more than two groups.

## Trying It Out

```{r}
anovaTest = aov(mpg ~ as.factor(gear), data = mtcars, projections = TRUE)
summary(anovaTest)
coefficients(anovaTest)
```


Remember:

\(F = \frac {explained\,variance} {unexplained\,variance}\)

## Contrasts

One interesting "advantage" of an ANOVA is that it gives you the ability to plan contrasts.

- You can do the same thing in a linear model!

Contrasts allow you to specify how to compare the different groups.

>-  Treatment/simple contrasts

>-  Helmert

>-  Sum


# What To Do?

## Which Is The Appropriate Method?

Hopefully, we can see that these are all *essentially* identical.

We need to think about what exactly we are doing:

- Are we predicting something?

- Are we concerned about group differences?

- Do we want to be limited?

- Are we doing experimental work?