---
title: "glm2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Factor Variables

Let's see if there might be anything interesting going on with the region variable:

```{r}
library(dplyr)

happy %>% 
  group_by(Region) %>% 
  summarize(meanHappy = mean(Happiness.Score))
```


```{r}
happy = read.csv("http://nd.edu/~sberry5/data/happy.csv")

factorTest = lm(Happiness.Score ~ Region, data = happy)
```

These are called treatment contrasts. If you want to change the treatment, try something like the following:

```{r}
factorTest2 = lm(Happiness.Score ~ relevel(happy$Region, ref = "North America"), 
                 data = happy)
```


# Interactions

```{r}
twoVars = lm(Happiness.Score ~ Freedom + Health..Life.Expectancy., data = happy)

summary(twoVars)
```


Let's explore interactions (moderation to some). 

```{r}
intMod = lm(Happiness.Score ~ Freedom * Health..Life.Expectancy., data = happy)

summary(intMod)
```


The interpretation of our main effects don't really change. We see, though, that our interaction has some lovely stars -- this requires interpretation. 

The effect of Freedom on Happiness increases by 7 units for every unit increase in life expectancy. If life expectancy is at 0, the slope between freedom and happiness -2.07. 

# Prediction


# T-tests

## What Are They Good For

You can use a *t*-test to test differences between two groups.

There are two general forms of the *t*-test:
  
- Independent

- Paired

## Our Focus

We are going to focus mostly on comparing independent samples.

Unless you are going to be doing experimental work, you will probably not need to use paired tests.

Furthermore, you probably won't ever really need to compare a sample to the population (requires you to know $\mu$)

## Tails

Like many other tests, the *t*-test can be tested with either one tail or two tails.

Alternative hypotheses can be any one of the following:

- $\neq$

- $>$

- $<$


What is the difference?

>-  Do you want to look like you know what you are doing or not?

## One Or Two

In all seriousness, let's consider the following plot:
  
  ```{r, eval = TRUE}
hist(rnorm(100000))
```


## Let's Give It A Try

```{r}
t.test(mtcars$mpg ~ mtcars$am, alternative = "two.sided")
```

Try it with different values for alternative and with var.equal = TRUE


## Multiple Tests

When conducting *t*-tests, you will often encounter the need to do a lot of them.

This leads to inflated rates of Type I errors

Many corrections exist:
  
- Bonferroni

- Tukey

- Scheffé

# Analysis Of Variance

## ANOVA

ANOVA is a lot like a *t*-test, but you can have more than two groups.

## Trying It Out

```{r}
anovaTest = aov(mpg ~ as.factor(gear), data = mtcars, projections = TRUE)
summary(anovaTest)
coefficients(anovaTest)
```


Remember:
  
  \(F = \frac {explained\,variance} {unexplained\,variance}\)

## Contrasts

One interesting "advantage" of an ANOVA is that it gives you the ability to plan contrasts.

- You can do the same thing in a linear model!
  
  Contrasts allow you to specify how to compare the different groups.

>-  Treatment/simple contrasts

>-  Helmert

>-  Sum


# What To Do?

## Which Is The Appropriate Method?

Hopefully, we can see that these are all *essentially* identical.

We need to think about what exactly we are doing:
  
  - Are we predicting something?
  
  - Are we concerned about group differences?
  
  - Do we want to be limited?
  
  - Are we doing experimental work?
  
  
  
## Rules Of Thumb

> - How many people do you need?

> - 20 records per predictor...

> - Not happening here!

> - We will use a more appropriate approach.

## Power Analysis

Do you want to melt most people's brains?

>-  Don't use rules of thumb!

>-  Instead of trusting outdated advice, use actual science to determine how many people you need to find if a difference exists.

## Power Analysis

We need three of the following parameters:

>-  Effect size

>-  Sample size

>-  Significance level

>-  Power

>-  We **should** always be doing this *a priori*

>-  Sometimes, it is fun to be a "statistical coroner"

## Effect Sizes

We won't get too deep here, just know that it is a way to determine the difference between groups.

- It's an improvement of *p*-values alone.

Cohen's *d* is likely the most common.

Let theory be your guide.

> - Be realistic!

## Power

Power is ability to detect an effect.

- In NHST words, we are trying to determine if we correctly reject the null hypothesis.

- Type I errors: Reject a true $H_{o}$ (false positive)

- Type II errors: Reject a false $H_{o}$ (false negative)

>- Which is more dangerous?

## Putting It All Together

Let's use the <span class="func">pwr</span> package.

```{r}
library(pwr)

pwr.f2.test(u = 1, v = NULL, f2 = .05, power = .8)

```

In the function: 

- u is the numerator df (*k* - 1)

- v is the denominator df (*n* - *k*) 

- f2 is signficance level

- \(\Pi = 1 -\beta\)

- \(\beta = Type\,II_{prob}\)


## Your Turn!

Use various values to do an *a priori* power analyses.

How does the proposed sample size change as the number of predictors goes up?

What if you tweak the significance level?

What about power?


## Different Test, Different Power Tests

We just did a test for a linear regression model.

Here is one for a *t*-test:

```{r}
tPower = pwr.t.test(n = NULL, d = 0.1, power = 0.8, 
                    type= "two.sample", alternative = "greater")

plot(tPower)
```


## Congratulations

>-  Your knowledge now far exceeds most professionals!
  