---
title: "Learning R"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: zenburn
    css: documentCSS.css
    center: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, comment = "")
```

## Why We Start At R

```{r, echo = FALSE, eval = TRUE}
fortunes::fortune(which = 52)
```


## Words Of Wisdom

```{r, echo = FALSE, eval = TRUE}
fortunes::fortune(which = 108)

fortunes::fortune(which = 323)
```


## More

```{r, echo = FALSE, eval = TRUE}
fortunes::fortune(which = 370)
```


## Hard To Argue

```{r, echo = FALSE, eval = TRUE}
fortunes::fortune(which = 168)
```

## What You Need To Know

R is an object-oriented statistical programming language.

```{r}
assortedNumbers = 15:30
```

```{r}
assortedNumbers[5]
```

R is vectorized.

```{r}
(assortedNumbers + 20) * 5
```

Others would need a loop...

```{r}
res = 1:length(assortedNumbers)

for(i in 1:length(assortedNumbers)){
  res[i] = (assortedNumbers[i] + 20) * 5
}
```


## Reading Data

There are more than a few ways to read data into R:

- <span class="func">read.csv()</span> -- Most common

- <span class="func">read.table()</span>

- <span class="func">read.delim()</span>

Sometimes people will give you data in special forms (.sav, .dta, .sas7bdat, etc.) and this is where haven and readxl have you covered.

  - You probably should not associate with folks who send you such files -- they are not to be trusted.

## Reading Data

Try this:

```{r}
testData = read.csv("http://www3.nd.edu/~sberry5/bball.csv")
```


## Getting Packages

R comes with several great packages.

- <span class="pack">base</span>, by itself, is more powerful than most other programs

There is also a massive eco-system around R and CRAN is the official repository.

```{r, eval = FALSE}
install.packages("dplyr")

library(dplyr)
```


## Object Types

At the data level, we are usually dealing with data frames or a matrix.

A data frame contains mixed data.

A matrix is all the same.

We often encounter lists, but we will save that for later.

## R Is So Hard...(Followed By Tears)

Sure it is, but so was walking before you knew how to do it.

This is a linear model:

```{r}
testModel = lm(mpg ~ hp + wt, data = mtcars)
```

Compare this simple line of code to that of a graphical interface and we can talk about which is faster and easier to follow.

And a summary:

```{r}
summary(mtcars)
```

##

### Python

```{r, eval = FALSE}
from sklearn import datasets, linear_model

regr = linear_model.LinearRegression()

# Two steps to get X and Y

regr.fit(X, Y)
```


### SAS

```{r, eval = FALSE}
proc reg;
  model mpg=hp wt;
run;
```


### SPSS
```{r, eval = FALSE}
regression/dependent mpg/enter hp wt.
```

### Stata

```{r, eval = FALSE}
regress mpg hp wt
```



## Handy Functions

We can look at the structure of our data with <span class="func">str()</span>.

All of the column names can be checked with <span class="func">names()</span>.

Getting a table view is easy with <span class="func">View()</span>.

The previously-seen <span class="func">summary()</span> can summarize your data and other objects.

You will see <span class="func">c()</span> all the time; it combines things into a vector.

The <span class="func">$</span> is one way to extract variables from a data frame.

- dat[, 'varName'] is another.

You might make very good friends with <span class="func">sink()</span>.

## A Little Less Conversation...

Now it is your turn!

Use the built-in mtcars data and inspect it.

## Helpful RStudio Hints

F1 *should* show the help files for any function.

F2 shows the function code.

Ctrl + Enter runs code.

Ctrl + Shift + P runs whatever you ran before, but with changes.

Ctrl + I indents your code.

Shift + Alt + K for everything

TAB...TAB...TAB


## Creating Your Own Functions

"If you are going to do something more than twice, just write a function." -- Michael Clark

R, unlike many others, makes function writing far too easy.

```{r}
myFirstFunction = function(x) {
  resSum = sum(x)
  res = resSum / length(x)
  return(res)
}
```

## Power Is Good...

As you continue your pursuit of statistical excellence, you will write functions.

```{r}
really = function(x){
  apply(x, 2, function(x) {
    sum(x) / length(x)
  })
}
```


## But Sometimes Painful

There was no reason to write such a function, when R already has one.

```{r}
colMeans()
```

It is a total "noob" mistake that even the most well-known researchers make.

- Don't feel bad when you do it.

## Combining Functions

You can *almost* infinitely chain functions with R.

- This is assuming that they are things that can chain without error.

```{r}
log(colMeans(mtcars[which(mtcars$cyl == 6), ]))
```

- Not that this particular bit of code is something you would ever need to do.



## A Modern Approach

We just saw a little bit of <span class="pack">base</span>.

- We all loved it, I know.

But now, let's check out what the <span class="pack">dplyr</span> package is all about.


```{r}
library(dplyr)
?dplyr
```

## dplyr

The <span class="pack">dplyr</span> package is part of the *tidyverse*.

At its core, it is providing a "grammar for data manipulation".

It utilizes a piping function (<span class="func">%>%</span>) to chain statements.

- The <span class="func">%>%</span> comes from <span class="pack">magrittr</span>

## Inside Out

You probably remember this:

```{r}
log(colMeans(mtcars[which(mtcars$cyl == 6), ]))
```

Again, we read this from the inside out.

## Maybe More Sense?

Contrast with the following:

```{r}
library(dplyr)

mtcars %>% 
  filter(cyl == 6) %>% 
  colMeans() %>% 
  log()
```

## Some Other Assorted Fun

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarize(n = n(), 
            meanMPG = mean(mpg), 
            sdMPG = sd(mpg))
```

```{r}
mtcars %>% 
  group_by(am) %>% 
  mutate(wtRaw = wt * 1000) %>% 
  select(-wt) %>% 
  summarize_all(mean)
```


## I Have Talked Long Enough

Engage in some <span class="pack">dplyr</span> fun for yourself.

Let's keep using the mtcars data and do the following:

1.  Pick a grouping variable(s) (am, cyl, gears) -- <span class="func">group_by</span>
2.  Create a new variable -- <span class="func">mutate</span>
3.  Create some type of summary