---
title: "Lab 2"
description: |
  Images To Data
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Extracting Data From Images (A Handy Aside)

Something that you might have to do at some point. It would take some work, but can definitely be done. 

NOTE: You do not need to do anything with this -- it is purely for demonstration.

```{r}
library(dplyr)
library(ggplot2)
library(magick)

sotuImage <- image_read("unstructured/notes/reading-level-state-of-the-union.png")

imageOut <- image_channel(sotuImage, "saturation") %>% 
  image_threshold("black", "05%") %>% 
  image_negate()  

imageData <- image_data(imageOut)[1, , ] %>%
  as.data.frame() %>% 
  mutate(Row = 1:nrow(.)) %>%
  select(Row, everything()) %>%
  mutate_all(as.character) %>%
  tidyr::pivot_longer(cols = starts_with("V")) %>%
  mutate(name = as.numeric(gsub("V", "", name)),
         Row = as.numeric(Row),
         value = ifelse(value == "00", NA, 1)) %>%
  filter(!is.na(value))

ggplot(imageData, aes(Row, name)) +
  geom_point()
```

# Reading PDF To Text

Getting data out of a pdf can seem like a daunting task. The `pdftools` and `tabulizer` packages can come to the rescue whenever we are presented with pdf. While `tabulizer` is best used for extracting tables out of pdf files, `pdftools` can handle most text-based situations.

## Trying Tabulizer

Just as an exercise, try to use `tabulizer` to read the tables in the St. Joe County file. Do note that this can cause weird issues for Macs -- it should not be too big of a hassle on a Windows machine. This is here so that you can see it; I do not want you wasting a ton of time trying to troubleshoot this (especially if you are using a Mac). 

```{r}
dyn.load('/Library/Java/JavaVirtualMachines/openjdk-13.0.1.jdk/Contents/Home/lib/server/libjvm.dylib')

library(tabulizer)

tabulizer::extract_tables("https://www.in.gov/dlgf/files/PropertyTaxManualCodeLists.pdf")
```

## Extracting Text

What should offer more success, though, is extracting text from the following two pdf files:

`Trial-Memorandum-of-President-Donald-J.-Trump.pdf`

`in_re_president_trump_house_impeachment_trial_brief_and_sof_1.18.20.pdf`

DO NOT READ THESE! Instead read both of these files in and produce some text descriptives (e.g., word frequency tables, wordclouds). Try to determine the goal of each document and maybe even the composing group.

```{r}
library(magick)
library(tesseract)
library(pdftools)

writtenstatement <- pdf_text("/Users/sethberry/Downloads/Trial-Memorandum-of-President-Donald-J.-Trump.pdf")
```

After completing your text exploration, create a three-slide story using any R-based slides.
