---
title: "Text Analysis"
description: |
  Lab 2
output:
  radix::radix_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Expanded Functions of Topic Models

Topic models can be extended to include covariates. If we add a variable to the prevalence argument, we are examining how another variable might predict the probability of a topic occuring within a document.


```{r}
library(dplyr)

library(stm)

library(stringr)

library(tm)

library(textstem)

load("C:/Users/sberry5/Documents/teaching/courses/unstructured/data/allLyricsDF.RData")

load("C:/Users/sberry5/Documents/teaching/courses/unstructured/data/countryTop50.RData")

allTop50 = allTop50 %>% 
  group_by(song) %>% 
  slice(1)

hardRemove = c(34, 78, 91, 94, 107, 136, 150, 202, 210, 213, 222, 229, 232, 
               239, 245, 248, 268, 276, 277, 284, 285, 291, 309, 310, 356, 
               403, 419, 424, 425, 460, 467, 482, 490, 497, 519, 532, 545, 
               555, 585, 586, 616, 639, 642, 648, 653, 654, 658, 682, 704, 
               705, 719, 722, 760, 780, 794, 795, 802, 832, 836, 851, 867, 
               872, 878, 891, 921, 924, 931, 944, 972, 985, 995, 1026, 1029, 
               1032, 1041, 1042, 1063, 1101, 1102, 1111, 1115, 1121, 1124,
               1126, 1128, 1144, 1145, 1158, 1166, 1172, 1175, 1186, 1202, 
               1204, 1209, 1211, 1215, 1218, 1225, 1240, 1251, 1256, 1257, 
               1272, 1287, 1311, 1316, 1326, 1327, 1342, 1361, 1421, 1462, 
               1493, 1534, 1584, 1609, 1627, 1629, 1635, 1638, 1649, 1667, 
               1676, 1679, 1690, 1696, 1707, 1710, 1717, 1720, 1755, 1758, 
               1760, 1780, 1786, 1788, 1798, 1805, 1847, 1855, 1858, 1861, 
               1866, 1882, 1895, 1918, 1929, 1946, 1952, 1994, 2000, 2011, 
               2017, 2018, 2019, 2027, 2032, 2039, 2055, 2069, 2143, 2164, 
               2193, 2200, 2201, 2204, 2246, 2254, 2285, 2303, 2318, 2334,
               2347, 2373, 2383, 2395, 2397, 2401, 2402, 2406, 2434, 2443,
               2459, 2469, 2472, 2475, 2499, 2507, 2527, 2534, 2540, 2563,
               2564, 2566, 2570, 2572, 2597, 2604, 2632, 2638, 2655, 2661,
               2666, 2685, 2696, 2719, 2721, 2724, 2729, 2736, 2740, 2754,
               2764, 2767, 2768, 2771, 2803, 2843, 2877, 2903, 2905, 2913, 
               2944, 2951, 2969, 2972, 2978, 2995, 3016, 3041, 3043, 3048, 3092)

cleanLyrics = allLyricsDF[-c(hardRemove)] %>%
  filter(warningIndicator == 0) %>% 
  dplyr::select(lyrics, returnedArtistName, returnedSong) %>%
  mutate(text = as.character(lyrics), 
         text = str_replace_all(text, "\n", " "),   
         text = str_replace_all(text, "(\\[.*?\\])", ""),
         text = str_squish(text), 
         text = gsub("([a-z])([A-Z])", "\\1 \\2", text), 
         text = tolower(text), 
         text = removeWords(text, c("’", stopwords(kind = "en"))), 
         text = removePunctuation(text), 
         text = removeNumbers(text),
         text = lemmatize_strings(text), 
         doc_id = returnedSong, 
         author = returnedArtistName) %>% 
  select(doc_id, text, author)

cleanLyrics = left_join(cleanLyrics, allTop50, by = c("doc_id" = "song"))

# To make things easier on the model side, we can just drop our NAs for now.

load("C:/Users/sberry5/Documents/teaching/courses/unstructured/data/countryTop50.RData")

cleanLyrics = na.omit(cleanLyrics)

cleanLyrics = cleanLyrics %>% 
  mutate(year = lubridate::year(date))

lyricText = textProcessor(documents = cleanLyrics$text, 
                          metadata = cleanLyrics, 
                          stem = FALSE)

lyricPrep = prepDocuments(documents = lyricText$documents, 
                               vocab = lyricText$vocab,
                               meta = lyricText$meta)

set.seed(1001)

topicPredictor = stm(documents = lyricPrep$documents, 
             vocab = lyricPrep$vocab, prevalence = ~ year,
             data = lyricPrep$meta, K = 5, verbose = FALSE, seed = 1001)

# If significant, overdispersion is a problem and more topics might be needed!

checkResiduals(topicPredictor, documents = lyricPrep$documents)

labelTopics(topicPredictor)
```

Let's look at the correlations between topics:

```{r}
plot.topicCorr(topicCorr(topicPredictor))
```

Not terribly surprising.


With our new model, we can actually test the effect of year on topic.

```{r}
set.seed(1001)

yearEffect = estimateEffect(1:5 ~ year, stmobj = topicPredictor, 
               metadata = lyricPrep$meta)

summary(yearEffect, topics = c(1:5))

plot.estimateEffect(yearEffect, "year", method = "continuous", 
                    model = topicPredictor, topics = 1, labeltype = "frex")

plot.estimateEffect(yearEffect, "year", method = "continuous", 
                    model = topicPredictor, topics = 2, labeltype = "frex")

plot.estimateEffect(yearEffect, "year", method = "continuous", 
                    model = topicPredictor, topics = 3, labeltype = "frex")

plot.estimateEffect(yearEffect, "year", method = "continuous", 
                    model = topicPredictor, topics = 4, labeltype = "frex")

plot.estimateEffect(yearEffect, "year", method = "continuous", 
                    model = topicPredictor, topics = 5, labeltype = "frex")
```

Absolutely amazing! As the years go by, we see an increase in party country and nostalgia, and a sharp decrease in sad love! I hate to say it, but maybe my sister-in-law is right -- country is changing.

We can add additional information to our models: content. What we get here is a better understanding about how people use language within a topic differently.

It should be noted that this will be a completely different model than what we had before. For these content-based models, we are saying that there is some difference between our content covariate with how words are used in topics, so our model will try to find the base words and then which words are used for the different levels of our content.

```{r, echo=FALSE}
load("C:/Users/sberry5/Documents/teaching/courses/unstructured/data/topicContent.RData")
```


```{r, eval = FALSE}
topicContent = stm(documents = lyricPrep$documents, 
             vocab = lyricPrep$vocab, content = ~ year,
             data = lyricPrep$meta, K = 5, verbose = FALSE, seed = 1001)
```

```{r}
labelTopics(topicContent)
```


## The Data

The data that we will be using comes from review data that some of us have already seen -- spectator feedback. You will find it in the data folder (called *feedbackData.RData*) and it contains only 3 variables: the sport, the feedback, and an overall satisfaction score.

## The Objectives

As a preliminary pass through the data, you might want to conduct a regular topic model (after doing any bit of text cleaning, of course). Try various values of *k* and make a decision based upon the output of `searchK`. After making your decision, run your topic model with your decided *k* value. Explore the results of your analyses and determine if you can create a coherent story for each of the topics.

After running the regular topic model, take some time to explore if any of the additional variables might be interesting to include as a covariate. 