---
title: "Classification"
description: |
  Lab 3
output:
  radix::radix_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = FALSE)
```

# Reading Tabular Data From PDF

Last week, we saw how to read text from an image. While this is certainly helpful, we saw that it was going to retain a line-by-line strucutre for that data. This would work well for anything with text, but it would not do any good with tabular data. 

<aside>
You can read text from a pdf with `pdftools`.
</aside>

If we want to (easily) retain some type of table strucutre from a pdf, we can use the tabulizer package to extract it out. You can pass along a pdf and pull any tables out. You can also do this interactively. 

```{r}
library(tabulizer)

testData = tabulizer::extract_tables("C://Users/sberry5/Documents/teaching/courses/unstructured/data/tableData.pdf")
```

Once your data is in, you will be faced with a list structure. Converting it should be simple from there.

```{r}
testData = as.data.frame(testData)

head(testData)
```

# Your Task

We have a whole slew of documents in a folder called *lab3Data*. You will find a great many .png files of text and a .pdf file with some information on it about the people. 

The first step would be reading in all of the text and getting it into a data structure. 

```{r}
library(tesseract)

doc1_1 = ocr("C://Users/sberry5/Documents/teaching/courses/unstructured/data/doc1_Page_1.png")

doc1_2 = ocr("C://Users/sberry5/Documents/teaching/courses/unstructured/data/doc1_Page_2.png")

# Hint -- you can get these together with the paste functiton and the collapse argument!
```

Remember, you will want to see if the image needs cleaned before getting onto any analyses.

Once we have all of our data read, cleaned, and organized, you might give a whirl to some form of classification. While we certainly do not have a ton of text to deal with here, it would be good to explore (you can definitely stretch this out to more data if you split on sentences!). 