---
title: "Classification"
description: |
  Lab 3
output:
  radix::radix_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = FALSE)
```

# Step 1 -- Reading Tabular Data From PDF.

Last week, we saw how to read text from a pdf or image. While this is certainly helpful, we saw that it was going to retain a line-by-line strucutre for that data. This would work well for anything with text, but it would not do any good with tabular data. 
If we want to (easily) retain some type of table strucutre from a pdf, we can use the tabulizer package to extract it out. You can pass along a pdf and pull any tables out. You can also do this interactively. 

```{r}
library(tabulizer)

testData = tabulizer::extract_tables("D:/projects/courses/unstructured/data/tableData.pdf")
```


Once your data is in, you will be faced with a list structure.

```{r}
testData = as.data.frame(testData)
```

# Step 2 -- Read In Text Data

# Step 3 -- Read In Table

# Step 4 -- Join And Classify

