---
title: "Text Analysis"
description: |
  Sentiment Analysis
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Processing Text

Before we can even begin to dive into analyzing text, we must first process the text. Processing text involves several steps that will be combined in various ways, depending on what we are trying to accomplish.

### Stemming

Tense aside, are chewed, chew, and chewing the same thing? Yes, but what if we compare the actual strings? On a string comparison side, are they the same? No. We have a string with 6, 4, and 7 characters, respectively.

What if we remove the suffixes, "ed" and "ing" -- we are left with three instances of "chew"? Now we have something that is equivalent in meaning and in a string sense. This is the goal of stemming.  

Let's take a look to see how this works (you will need to install tm and SnowballC first):

```{r}
chewStrings = c("chew", "chewing", "chewed", "chewer")

tm::stemDocument(chewStrings)
```


We got exactly what we expected, right? You might have noticed that "chewer" did not get stemmed. Do you have any idea why? Let's think through it together. "Chew", "chewing", and "chewed" are all verbs related to the act of chewinging. "Chewer", on the other hand, is a person who chews -- it is a noun. Martin Porter's stemming algorithm works incredibly well!

Hopefully, this makes conceptual sense; however, we also need to understand why we need to do it. In a great many text-based methods, we are going to create a matrix that keeps track of every term (i.e., word) in every document -- this is know as a document-term matrix. If we know that "chew", "chewing", and "chewed" all refer to the same thing, we want it just represented once within our document-term matrix.

Shall we take a look?

```{r}
library(tm)

documents = c("I like to chew", 
              "I have chewed my whole life", 
              "Chewing and stomping through the fields", 
              "I am a chewer")

documentsCorp = tm::SimpleCorpus(VectorSource(documents))

documentsDTM = DocumentTermMatrix(documentsCorp)

inspect(documentsDTM)
```

We can see that without stemming, we have 9 terms (things like "I", "a", and "to" get removed automatically). Let's do some stemming now:

```{r}
documentsStemmed = stemDocument(documents)

documentsStemmed
```

And now the document-term matrix:

```{r}
stemmedDocCorp = tm::SimpleCorpus(VectorSource(documentsStemmed))

stemmedDocDTM = DocumentTermMatrix(stemmedDocCorp)

inspect(stemmedDocDTM)
```

If we are trying to find documents that are covering similar content or talking about similar things, this document-term matrix will help to draw better conclusions, because it is clear that the first three documents are talking about the act of chewing and this document-term matrix reflects that.

### Lemmatization

Stemming is often enough (and most modern stemmers work pretty well on their own). Still, stemming is slightly more akin to amputating an arm with a battle ax -- it works, but it is brute force. Lemmatization is a more sophisticated approach. You might have already guessed that lemmatization will find the *lemma* of a word and since you likely know about morphology, you already know that the lemma of a word is its canonical form. A group of words that form the same idea are called a lexeme (am, be, are are all within the same lexeme). Generally, the smallest form of the word is chosen as the lemma. This is a really interesting area of linguistics, but we don't need to dive fully in.  

Instead, let's see it in action.

If we compare some "chewing" stuff on stemming and lemmatization, we can see what we get:

```{r}
library(textstem)

chewStrings = c("chew", "chewing", "chewed", "chewer")

stem_words(chewStrings)

lemmatize_words(chewStrings)
```

Absolutely nothing different. Both stemming and lemmatizing will perform the same task. The act of chewing is comprised of a past, present, and future tense, and chew is the lemma; chewer is still seen as something else entirely.

But let's take a look at something different. If we have a string of the most lovely words, what might happen?

```{r}
lovelyString = c("lovely", "lovelier", "loveliest")

stem_words(lovelyString)
```

That is about as close to "bigly" nonsense as we could possibly get without going into Dr. Suess mode. 

But if we try lemmatization:

```{r}
lemmatize_words(lovelyString)
```

We get something that starts to make sense. Now, let's try these on some actual chunks of text and see what happens.

```{r}
# This data is in the "data" folder on Sakai!

load("C:/Users/sberry5/Documents/teaching/courses/unstructured/data/allLyricsDF.RData")

sampleLyrics = allLyricsDF[40, ]

sampleLyrics$lyrics
```

Of course, we will need to do some cleaning on our text first:

<aside>
As we work with this lyric stuff, you will need to make sure that you clean it before playing with it. The code that we saw last week should take care of it...maybe!
</aside>

```{r}
library(dplyr)

library(stringr)

cleanLyrics = sampleLyrics$lyrics %>% 
  str_replace_all(., "\n", " ") %>% 
  str_replace_all(., "\\[[A-Za-z]+\\s*[0-9]*]", "") %>%
  str_squish(.) %>% 
  gsub("([a-z])([A-Z])", "\\1 \\2", .)
```

We have to try the obligatory stemming:

```{r}
stem_strings(cleanLyrics)
```

And now the lemmatized version:

```{r}
lemmatize_strings(cleanLyrics)
```

Here is something very interesting:

```{r}
microbenchmark::microbenchmark(stem_strings(cleanLyrics), 
                               lemmatize_strings(cleanLyrics))
```

What is the point here? This song has a little over 400 words in it. Stemming, over 100 runs, took on average 1.3 milliseconds, while lemmatizing took 3.6. We are just talking milliseconds, so this is almost to the point where we would not notice; but, if we did this over an entire corpus, we would definitely notice the time.

The question, then, is what do you decide to do. For my money, lemmatization does a better job and getting words down to their actual meaning.

### Stop Words

Some words do us very little good: articles, prepistions, and very high frequency words. These are all words that need to be removed. Fortunately, you don't have to do this on your own -- a great many dictionaries exist that contain words ready for removal.

```{r}
tm::stopwords("en")
```

Removing stopwords takes little effort!

```{r}
documents = c("I like to chew.", 
              "I am stompin and chewing.", 
              "Chewing is in my blood.", 
              "I am a chewer")

tm::removeWords(documents, words = stopwords("en"))
```

We can even include custom stopwords:

```{r}
tm::removeWords(documents, words = c("blood", stopwords("en")))
```

There are many different stopword lists out there, so you might want to poke around just a little bit to find something that will suit the needs of a particular project. 

```{r}
library(stopwords)
```


## Text Processing Tools

There are several R packages that will help us process text. The tm package is popular and automates most of our work. You already saw how we use the stemming and stopword removal functions, but tm is full of fun stuff and allows for one pass text processing.

```{r, eval = FALSE}

documents = c("I like to chew.", 
              "I am stompin and chewing.", 
              "Chewing is in my blood.", 
              "I am a chewer")

documentCorp = SimpleCorpus(VectorSource(documents))

stopWordRemoval = function(x) {
  removeWords(x, stopwords("en"))
}

textPrepFunctions = list(tolower,
                         removePunctuation,
                         lemmatize_strings,
                         stopWordRemoval,
                         removeNumbers,
                         stripWhitespace)

documentCorp = tm_map(documentCorp, FUN = tm_reduce, tmFuns = textPrepFunctions)

documentCorp[1][[1]]$content
```


Once you get your text tidied up (or even before), you can produce some visualizations!

```{r}
library(tidytext)

library(wordcloud2)

allLyricsDF %>%
  filter(warningIndicator == 0) %>% 
  dplyr::select(lyrics, returnedArtistName) %>%
  mutate(lyrics = as.character(lyrics), 
         lyrics = str_replace_all(lyrics, "\n", " "),   
         lyrics = str_replace_all(lyrics, "\\[[A-Za-z]+\\s*[0-9]*]", ""), 
         lyrics = str_squish(lyrics), 
         lyrics = gsub("([a-z])([A-Z])", "\\1 \\2", lyrics)) %>%
  unnest_tokens(word, lyrics) %>% 
  anti_join(stop_words) %>% 
  count(word, sort = TRUE) %>% 
  filter(n > 25) %>% 
  na.omit() %>% 
  wordcloud2(shape = "cardioid")
```


## Sentiment Analysis

Sentiment analysis is commonly used when we want to know the general *feelings* of what someone has written or said. Sentiment analysis is commonly seen applied to Twitter and other social media posts, but we can use it anywhere where people have written/said something (product reviews, song lyrics, final statements).

Sentiment can take many different forms: positive/negative affect, emotional states, and even financial contexts.

Let's take a peak at some simple sentiment analysis.

### Simple Sentiment

Let's consider the following statements:


```{r}
library(tidytext)

statement = "I dislike beer, but I really love the shine."

tokens = data_frame(text = statement) %>% 
  unnest_tokens(tbl = ., output = word, input = text)

tokens
```

Now, we can compare the tokens within our statement to some pre-defined dictionary of positive and negative words.

```{r}
library(tidyr)

tokens %>%
  inner_join(get_sentiments("bing")) %>% 
  count(sentiment) %>% 
  spread(sentiment, n, fill = 0) %>% 
  mutate(sentiment = positive - negative)
```

When we use Bing's dictionary, we see that we get one positive word (love) and negative word (dislike) with a neutral overall sentiment (a sentiment of 0 would indicate neutrality, while anything above 0 has an increasing amount of positivity and anything below 0 has an increasing amount of negativity).

Do you think that disklike and love are of the same magnitude? If I had to make a wild guess, I might say that love is stronger than dislike. Let's switch out our sentiment library to get something with a little better notion of polarity magnitute.

```{r}
tokens %>%
  inner_join(get_sentiments("afinn"))
```

Now this looks a bit more interesting! "Love" has a stronger positive polarity than "dislike" has negative polarity. So, we could guess that we would have some positive sentiment.

If we divide the sum of our word sentiments by the number of words within the dictionary, we should get an idea of our sentences overall sentiment.

```{r}
tokens %>%
  inner_join(get_sentiments("afinn")) %>% 
  summarize(n = nrow(.), sentSum = sum(score)) %>% 
  mutate(sentiment = sentSum / n)
```

Our sentiment of .5 tells us that our sentence is positive, even if only slightly so.

While these simple sentiment analyses provide some decent measures to the sentiment of our text, we are ignoring big chunks of our text by just counting keywords.

For example, it is probably fair to say that "really love" is stronger than just "love". We might want to switch over to some techniques that consider *n*-grams and other text features to calculate sentiment.

### Smarter Sentiment Analysis

When we use sentiment analysis that is aware of context, valence ("love" is stronger than "like"), modifiers (e.g., "really love"), and adversative statements ("but,...", "however,..."), we get a better idea about the real sentiment of the text.

We will use the *jockers* sentiment library, but many more available. Depending on your exact needs, there are some dictionaries designed for different applications. 

Before we engage in our whole sentiment analysis, let's take a look at a few things.

Here is the dictionary that *jockers* will use.

```{r}
lexicon::hash_sentiment_jockers
```

You might want to use View() to get a complete look at what is happening in there.

We should also take a peak at our valence shifters:

```{r}
lexicon::hash_valence_shifters
```

With all of that out of the way, let's get down to the matter at hand:

```{r}
library(sentimentr)

library(lexicon)

library(magrittr)

statement = "I dislike beer, but I really love the shine."

sentiment(statement, polarity_dt = lexicon::hash_sentiment_jockers)
```

We can see that we get a much stronger sentiment score when we include more information within the sentence. While the first part of our sentence starts out with a negative word (dislike has a sentiment value of -1), we have an adversarial "but" that will downweight whatever is in the initial phrase and then we will have the amplified (from "really") sentiment of "love" (with a weight of .75 in our dictionary).

With all of this together, we get a much better idea about the sentiment of our text.

### Back To The Music

While the text that we have used so far serves its purpose as an example quite well, we can always take a look at other written words.

```{r}
load(url("https://raw.githubusercontent.com/saberry/courses/master/hash_sentiment_vadar.RData"))

cleanLyrics = allLyricsDF %>%
  filter(warningIndicator == 0) %>% 
  dplyr::select(lyrics, returnedArtistName, returnedSong) %>%
  mutate(lyrics = as.character(lyrics), 
         lyrics = str_replace_all(lyrics, "\n", " "),   
         lyrics = str_replace_all(lyrics, "(\\[.*?\\])", ""), # look different?
         lyrics = str_squish(lyrics), 
         lyrics = gsub("([a-z])([A-Z])", "\\1 \\2", lyrics))

songSentiment = sentiment(get_sentences(cleanLyrics), 
          polarity_dt = hash_sentiment_vadar) %>% 
  group_by(returnedSong) %>% 
  summarize(meanSentiment = mean(sentiment))
```

Naturally, we would want to join those sentiment values up with our original data:

```{r}
cleanLyrics = left_join(cleanLyrics, songSentiment, by = "returnedSong")
```


From here, we have several choices in front of us. One, we could use those sentiment values within a model (e.g., we might want to predict charting position). Or, we could use them for some further exploration:

```{r}
library(DT)

sentimentBreaks = c(-1.7, -.5, 0, .5, 1.7)

breakColors = c('rgb(178,24,43)','rgb(239,138,98)','rgb(253,219,199)','rgb(209,229,240)','rgb(103,169,207)','rgb(33,102,172)')

datatable(cleanLyrics, rownames = FALSE, 
              options = list(pageLength = 15, escape = FALSE, 
                             columnDefs = list(list(targets = 1, visible = FALSE)))) %>% 
  formatStyle("lyrics", "meanSentiment", backgroundColor = styleInterval(sentimentBreaks, breakColors))
```


We can also do some checking over time:

```{r}
library(ggplot2)

load("C:/Users/sberry5/Documents/teaching/courses/unstructured/data/countryTop50.RData")

allTop50 = allTop50 %>% 
  group_by(song) %>% 
  slice(1)

cleanLyrics = left_join(cleanLyrics, allTop50, by = c("returnedSong" = "song"))

cleanLyrics %>% 
  group_by(date) %>% 
  na.omit() %>% 
  summarize(meanSentiment = mean(meanSentiment)) %>% 
  ggplot(., aes(date, meanSentiment)) + 
  geom_point() +
  theme_minimal()
```


That is pretty messy (but I am curious about that really happy month), so let's try something else:

```{r}
library(gganimate)

cleanLyrics %>% 
  mutate(year = lubridate::year(date), 
         month = lubridate::month(date)) %>% 
  group_by(year, month, date) %>% 
  na.omit() %>% 
  summarize(meanSentiment = mean(meanSentiment)) %>% 
  ggplot(., aes(as.factor(month), meanSentiment, color = meanSentiment)) + 
  geom_point() +
  scale_color_distiller(type = "div") +
  theme_minimal() +
  transition_states(year,
                    transition_length = length(1975:2018),
                    state_length = 3) +
  ggtitle('Year: {closest_state}')
```



```{r}
cleanLyrics %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year) %>% 
  na.omit() %>% 
  summarize(meanSentiment = mean(meanSentiment)) %>% 
  ggplot(., aes(year, meanSentiment, color = meanSentiment)) + 
  geom_point() +
  theme_minimal()
```


## Other Text Fun

Sentiment analysis is always a handy tool to have around. You might also want to explore other descriptive aspects of your text.

The koRpus package allows for all types of interesting types descriptives. There are a great number of readability and lexical diversity statistics (Fucks is likely my favorite).

We need to tokenize our text in a manner that will please koRpus.

### Embeddings

English is hard and seeing how words relate to other words can be tricky. Let's think through something weird:

$$\text{church} - \text{jesus} + \text{muhammad} = \text{mosque}$$

Or how about:

$$\text{Washing D.C.} - \text{America} + \text{Mexico} = \text{Mexico City}$$

What is the purpose here? Word embeddings start to break down how words can be different, but still deal in the same contextual space. Whether we are talking about a church or a mosque, we are just dealing with a place of worship. If we have two different text documents (one talking about mosques and one talking about churches), it would be nice to be able to recognize that they are largely talking about the same idea.

```{r}
library(text2vec)

horrorTxt <- rbind(f13, halloween, evilDead)

tokens <- space_tokenizer(horrorTxt)
# Create vocabulary. Terms will be unigrams (simple words).

it <- itoken(tokens, progressbar = FALSE)

vocab <- create_vocabulary(it)

vocab <- prune_vocabulary(vocab, term_count_min = 5L)

vectorizer <- vocab_vectorizer(vocab)

tcm <- create_tcm(it, vectorizer, skip_grams_window = 5L)

glove <- GlobalVectors$new(word_vectors_size = 50, vocabulary = vocab, x_max = 10)

glove_main <- glove$fit_transform(tcm, n_iter = 20)

glove_context <- glove$components

word_vectors <- glove_main + t(glove_context)

f13Vec <- word_vectors["halloween", , drop = FALSE] - 
  word_vectors["myers", , drop = FALSE] + 
  word_vectors["voorhees", , drop = FALSE]

cos_sim = sim2(x = word_vectors, y = f13Vec, method = "cosine", norm = "l2")

head(sort(cos_sim[,1], decreasing = TRUE), 5)
```

<aside>
If you looked at the Selenium code and noticed some of the code above, you might have seen some code behaving in a way that seems very similar to Python. Code like `glove$fit_transform()` is using the `R6` class. It is not a base R class, but is a package.
</aside>