---
title: "Homework 2"
author: "Your Name Here"
date: "11/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Bronze

### Part 1

Canopy Growth is trying to determine how to best utilize their product to make the most money. 

Current market prices (in Canada) would suggest each selling unit of oil will produce a revenue of \$40, while a selling unit of flower will yield \$7.50. 

To produce a selling unit of oil, 30 grams of flower is need; a selling unit of oil requires 3 grams. Furthermore, each unit of oil costs \$3.15 produce; a unit of flower costs \$.50 to produce. 

Currently, Canopy Growth only has 8960 grams available.

You have 3 things to report:

1.  What is the production split for oil and flower?

2.  What is the total revenue?

3.  What is the profit?

```{r}
library(linprog)

objectiveFunction <- c(40, 7.50)

names(objectiveFunction) <- c("oil", "flower")

constraintValues <- 8960

constraintMatrix <- rbind(c(30, 3))

linprog::solveLP(objectiveFunction, constraintValues, constraintMatrix, maximum = TRUE)
```

### Part 2

Given the market conditions that exist, Canopy Growth needs to produce at least 50 units of oil to satisify holiday demand. You will need to use the ROI package to specify lower and upper bounds on your objective functions.

You have 3 things to report:

1.  What is the production split for oil and flower, given our new bounds?

2.  What is the total revenue?

3.  What is the profit?

```{r}
library(ROI)

cgOP <- OP(objective = c(40, 7.50), 
           constraints = L_constraint(L = matrix(c(30, 3), nrow = 1, ncol = 2), 
                                      dir = "<=", 
                                      rhs = 8960), 
           maximum = TRUE, 
           bounds = list(upper = c(Inf, Inf), 
                         lower = c(50, 0)))

result <- ROI_solve(cgOP)

solution(result)
```


## Silver

Given your knowledge of linear programming, you have been asked to consult with a group of nuclear medicine doctors. They are curious about targeting cancerous cells in such a way that minimizes the harm to other cells. They are using external beam radiation, so there are two beams targeting a tumor.

In this regard, we have the following cell types:

- normal tissue (n)

- critical tissue(c; think important organ tissue)

- target tissue (t)

Each of these cell types requires a certain degree of average radiation exposure (measured in Kilorads) from the two beams. 

The following table has the requirements and the fraction of dose absorbed by each area for beams 1 and 2.

```{r}
data.frame(area = c("n", "c", "t", "tCenter"), 
           beam1 = c(.4, .3, .5, .6),
           beam2 = c(.5, .1, .5, .5), 
           restriction = c("minimize", "<= 2.7", "= 6", ">= 6"))
```


What is the optimal dose for beams 1 and 2?

## Gold

https://thalis.math.upatras.gr/~tsantas/DownLoadFiles/Hillier&Lieberman_7th-edition_Chapter3.pdf