---
title: "Homework 2"
author: "Your Name Here"
date: "11/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Bronze

### Part 1

Canopy Growth is trying to determine how to best utilize their product to make the most money. 

Current market prices (in Canada) would suggest each selling unit of oil will produce a revenue of \$40, while a selling unit of flower will yield \$7.50. 

To produce a selling unit of oil, 30 grams of flower is need; a selling unit of oil requires 3 grams. Furthermore, each unit of oil costs \$3.15 produce; a unit of flower costs \$.50 to produce. 

Currently, Canopy Growth only has 8960 grams available.

You have 3 things to report:

1.  What is the production split for oil and flower?

2.  What is the total revenue?

3.  What is the profit?

```{r}
library(linprog)

objectiveFunction <- c()

names(objectiveFunction) <- c("oil", "flower")

constraintValues <- 8960

constraintMatrix <- rbind(c())

linprog::solveLP(objectiveFunction, constraintValues, constraintMatrix, maximum = TRUE)
```

### Part 2

Given the market conditions that exist, Canopy Growth needs to produce at least 50 units of oil to satisify holiday demand (this is the lowerbound value for this variable). You will need to use the ROI package to specify lower and upper bounds on your objective functions.

You have 3 things to report:

1.  What is the production split for oil and flower, given our new bounds?

2.  What is the total revenue?

3.  What is the profit?

```{r}
library(ROI)

variableBounds <- V_bound(li = 1:2, ui = 1:2, lb = c(), ub = c(Inf, Inf))

cgOP <- OP(objective = c(40, 7.50), 
           constraints = L_constraint(L = matrix(c(30, 3), nrow = 1, ncol = 2), 
                                      dir = "<=", 
                                      rhs = 8960), 
           maximum = TRUE, 
           bounds = variableBounds)

result <- ROI_solve(cgOP)

solution(result)

result$message
```


## Silver

Given your knowledge of linear programming, you have been asked to consult with a group of nuclear medicine doctors. They are curious about targeting cancerous cells in such a way that minimizes the harm to other cells. They are using external beam radiation, so there are two beams targeting a tumor.

In this regard, we have the following cell types:

- normal tissue (n)

- critical tissue(c; think important organ tissue)

- target tissue (t)

Each of these cell types requires a certain degree of average radiation exposure (measured in Kilorads) from the two beams. 

The following table has the requirements and the fraction of dose absorbed by each area for beams 1 and 2.

```{r}
data.frame(area = c("n", "c", "t", "tCenter"), 
           beam1 = c(.4, .3, .5, .6),
           beam2 = c(.5, .1, .5, .5), 
           restriction = c("minimize", "<= 2.7", "= 6", ">= 6"))
```


What is the optimal dose for beams 1 and 2?

## Gold

A steel mill is trying to reduce emissions of three particular kinds of air pollutants.

The following are clean air standards for the facility (in millions of pounds):

```{r}
data.frame(pollutant = c("particulates", "sulfer oxides", "hydrocarbons"), 
           requiredReduction = c(60, 150, 125))
```

The steel mill has two major sources of these pollutants: blast furnaces and open-hearth furnances. To reduce the pollutants from these two furnace types, engineers are exploring using taller smokestacks, filters, and cleaner fuels.

The following are maximum estimated reduction rates (in millions of pounds) for various abatement methods:

```{r}
data.frame(pollutant = c("particulates", "sulfer oxides", "hydrocarbons"), 
           ts_blastFurnace = c(12, 35, 37), 
           ts_openHearth = c(9, 42, 53), 
           filter_blastFurnace = c(25, 18, 28), 
           filter_openHearth = c(20, 31, 24), 
           fuels_blastFurnace = c(17, 56, 29), 
           fuels_openHearth = c(13, 49, 20))
```

Each of these abatement methods can be used to any proportion.

The following table specifies the cost for fully-adopting each method (if the proportion is 1):

```{r}
data.frame(method = c("taller smokestacks", "filters", "fuels"), 
           blastFurnaces = c(8, 7, 11), 
           openHearth = c(10, 6, 9))
```


The ultimate goal of the steel mill is to minimize the money spent on each abatement method, while hitting the require reduction thresholds.

This creates the following linear program:

$$Minimize Z = 8_{x1} + 10_{x2} +7_{x3} +6_{x4} + 11_{x5} + 9_{x6}$$

**Subject to the following emission reduction constrains:**

$$12_{x1} + 9_{x2} + 25_{x3} + 20_{x4} + 17_{x5} + 13_{x6} \geq 60 \\
35_{x1} + 42_{x2} + 18_{x3} + 31_{x4} + 56_{x5} + 49_{x6} \geq 150 \\
37_{x1} + 53_{x2} + 28_{x3} + 24_{x4} + 29_{x5} + 20_{x6} \geq 125
$$

**Subject to the following technology limitation constrains:**

$$x_j \leq 1$$

What is each abatement methods optimal proportion?