---
title: "Mixed Models"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")
```

## Packages

We will need the following:

```{r, eval = FALSE}
install.packages(c("lme4", "lmerTest", "merTools"))
```


## Terminology

While the specifics of each model you have learned to this point might take some time to get our heads all the way around, the terminology has largely been pretty clear -- no more. You will hear "mixed models", "mixed effects models", "hierarchical linear models", "nested models", and/or "multilevel models"; these are all slight variations on a common theme. For the sake of our work here, we will keep it at mixed models. Within our mixed model, we have an additional source of cloudiness: fixed and random effects. The random effects don't pose much of an issue (we will define it later), but fixed effects have 4 different definitions depending upon whom you ask. For the sake of simplicity (again), we are going to consider fixed effects as an effect on the individual unit of analysis. This will all start to make sense once we take a look at the models.

## Standard Linear Model

For the sake of conceptual grounding, let's go back to our standard linear model:

```{r}
library(dplyr)

library(ggplot2)

healthData = readr::read_csv("https://www.nd.edu/~sberry5/data/healthViolationsDistances.csv")

healthData = healthData %>% 
  mutate(BORO = as.factor(.$BORO), 
         cuisine = as.factor(.$`CUISINE DESCRIPTION`), 
         distanceCentered = dohDistanceMeter - mean(dohDistanceMeter))

lmTest = lm(SCORE ~ distanceCentered, data = healthData)

ggplot(healthData, aes(SCORE, distanceCentered)) +
  geom_point() + 
  geom_smooth(method = "lm")

summary(lmTest)
```

We have our standard output here. As before, our intercept is the average score when there is zero distance between the restaurant and department of health and our coefficient for distance is telling us that for every mile increase in distance, we are increasing our score by some tiny amount. We know that we are ignoring some information within our model, namely the clustering that occurs based upon cuisine and/or borough. 


```{r}
ggplot(healthData, aes(SCORE, distanceCentered, group = BORO)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  facet_wrap( ~ BORO)
```


### Why Mixed Models

When we have repeated samples/measurement, mixed models will not overfit or underfit. They don't get bogged down by large groups and the smaller groups do not get buried by the larger groups.

In linear models, you likely learned about balance in *t*-tests; mixed models will attenuate the effects of group imbalance.

## Random Intercepts Model

Let's include borough in our model. We are not going to add it as another predictor, but we are going to include it as another level to our model. The **lme4** package will make this very easy:

```{r}
library(lme4)

riMod = lmer(SCORE ~ distanceCentered + (1|BORO), data = healthData)
```

Before we look at the summary for this model, let's get an idea about what is happening in the syntax. The first part of our formula should look familiar -- these are the global estimates (fixed effects) within our model and will behave exactly the same as our standard linear model. 

The next part in the parentheses is how we denote our random effect. Whenever you see a 1 included in a formula interface, we can be pretty comfortable that it is in reference to a intercept. The *|* specifies a grouping. With that information, we might be able to guess that we are specifying a random intercept for each borough. 

We should probably check out the summary:

```{r}
summary(riMod)
```

We have our standard output and we can see that our coefficients have changed from our standard linear model -- this change is purely due to the severe imbalance between our groups:

```{r}
healthData %>% 
  group_by(BORO) %>% 
  summarize(n())
```

When groups are largely balanced, we would find that our coefficients would be the same (or very close to it).

What should almost always change is our standard errors -- by integrating information about the groups, we are getting a better sense of how much uncertainty our model contains at the global average level.

We also see some additional information -- this is for our random effects. The standard deviation is telling us how much the score moves around based upon borough after getting the information from our fixed effects (i.e., the health score can move around nearly 1 whole point from boro alone). We can compare the standard deviation for BORO to the coefficient for distanceCentered -- Borough is contributing to more variability within Scores than distance. We can also add the variance components and divide by the random effects variance to get its variance account for.

```{r}
.9879 / (.9879 + 189.9733)
```

So while it might be doing more than what distance does, borough is not accounting for too much variance.

We can also plot simulated random effect ranges for each of the random effect groups. We want to pay attention to those that are highlighed with black (i.e., the range does not cross the red line at 0). 

```{r}
library(merTools)

plotREsim(REsim(riMod))

levels(healthData$BORO)
```

In examining the plot, we see that the random effect ranges for the Bronx and Staten Island have significant effects on health score.

Going back to the output, did you notice anything missing: *p*-values! Estimating *p*-values in a mixed model is exceedingly difficult because of varying group sizes, complete sample *n*, and how those relate to reference distributions. If you need something that will help, you can get confidence intervals in the same way that you would anything else:

```{r}
confint(riMod)
```


If you *really* want to see *p*-values, you can get them easily:

```{r}

riModP = lmerTest::lmer(SCORE ~ distanceCentered + (1|BORO), data = healthData)

summary(riModP)

```

**NOTE:** I would never load the lmerTest package, but would attach with colons!

Getting predicted values from our mixed model is no different then getting them from any other model:

```{r}
mixedPred = predict(riMod)

slimPred = predict(lmTest)

head(cbind(actual = healthData$SCORE, 
      mixed = mixedPred, 
      slim = slimPred), 20)

```

While there were cases where the standard linear model did a slightly better job, our mixed model generally did a better job (even if marginally so).

Let's add some more information to our model. As we dive into our data, we will notice that we also have cuisine groupings. We can add this additional grouping into our model:

```{r}
clusterMod = lmer(SCORE ~ distanceCentered + (1|cuisine) + (1|BORO), data = healthData)
```

This is often called a cross-classified model. 

```{r}
summary(clusterMod)
```

Let's look at our variances how we did earlier:

```{r}
# cuisine

2.90 / (2.90 + 1.867 + 186.933)

# BORO

1.867 / (1.867 + 2.90 + 186.933)
```


```{r}
plotREsim(REsim(clusterMod))
```


We should also see if our predictions improved:

```{r}
clusterPredict = predict(clusterMod)

head(cbind(actual = healthData$SCORE, 
           clustered = clusterPredict,
           mixed = mixedPred, 
           slim = slimPred), 20)
```

For many observations, our predictions definitely go tighter (many are still far off, though).

If we continue to look at our data (and with some knowledge about how NYC does health inspections), we will see that restaurants are rated yearly -- let's use this information in our model. We won't worry about distance anymore, because now we have a few competing hypotheses.

Let's do a bit of data processing first.

```{r}
healthDataGrouped = healthData %>% 
  tidyr::unite(col = nameLocation, DBA, BUILDING , remove = FALSE) %>% 
  group_by(nameLocation) %>%
  arrange(lubridate::mdy(`GRADE DATE`)) %>% 
  mutate(observation = 1:n())

timeReviewed = healthDataGrouped %>% 
  summarize(n = n()) %>% 
  filter(n > 10)

reviewedRest = healthDataGrouped[which(healthDataGrouped$nameLocation %in% 
                                         timeReviewed$nameLocation), ]
```


```{r}
observationMod = lmer(SCORE ~ observation + (1|nameLocation), data = reviewedRest)
```

In this model, we have a fixed effect for observation and we are allowing each location to have it's own random intercept. We have essentially created a model that will deal with the repeated measures for each of the locations.

```{r}
reviewedRest %>% 
  arrange(nameLocation, observation) %>% 
  head(., 350) %>% 
  ggplot(., aes(observation, SCORE, group = nameLocation)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap( ~ nameLocation) +
  theme_minimal()
```

```{r}
summary(observationMod)
```

Our fixed effect here would indicate that we have a slight increase in scores as our observations increase, but we can also see that scores will bounce around about 5 points on average by location alone.

```{r}
29.46 / (29.46 + 125.82)
```

We see that the location alone accounts for nearly 20% of the variance in health scores. 

## Hierarchical Models

Hierarchical are a slight variation on the models that we have just seen. In these models, we have groups nested within other groups. We know that we have a "BORO" variable, but a quick look at our data will show that we also have a DBA variable; this variable is giving the name of the restaurant.  One source of inquiry would be to inspect how various chain restaurants, nested within each boro, perform.

Let's check out how some chain restaurants do within the boroughs. 

```{r}
chainFood = healthDataGrouped %>% 
  filter(DBA == "BURGER KING" |
           DBA == "MCDONALD'S" | 
           DBA == "PIZZA HUT" |
           DBA == "SUBWAY")
```

The model set-up is just different enough to cause some potential confusion here. Within the parentheses, we have our intercept as before, but we are also saying that we are looking at the DBA groups within the BORO groups.

```{r}
hierMod = lme4::lmer(SCORE ~ observation + (1|BORO/DBA), 
                     data = chainFood)

summary(hierMod)
```

```{r}
plotREsim(REsim(hierMod))
```

```{r}

unique(hierMod@flist$BORO)
unique(hierMod@flist$`DBA:BORO`)
```

