---
title: "Week 2 Practice"
author: "BDS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
```

Many, if not all, organizations really care about employee retention -- replacing employees is a costly endeavor and one that organizations would rather not engage in with great regularity. To that end, many people involved in HR-flavored analytics are curious about what drives turnover -- is it pay, performance, interest, or something about the person?

Unfortunately, not all organizations have the capacity to explore such issues. If an organization came to you and asked about the causes of employee turnover, you might be able to help them; however, the level to which you could help them is going to be a function of the data at hand. Whenever we are looking at behaviors, we need to be sure that we actually have the behavior within the data. J

For example, could you provide any answers about turnover with the following data:

```{r}
data.frame(tenure = round(rnorm(500, mean = 10, sd = 2)), 
           salary = rnorm(500, mean = 45000, sd = 10000),
           last_review = sample(c("Developing", "Valued", "Top Performer"), 500, replace = TRUE)) %>% 
  summary()
```

There is nothing here about turnover, but we can see if there is any relationship between salary and tenure (I would propose that a longer tenure causes an increase in salary). We could also explore the relationship between review and salary (the better the review, the higher the salary -- this one is a bit tenuous, though, as yearly raise percentage would be better). 


What about the following data. What questions about turnover could be answered?

```{r}
data.frame(tenure = round(rnorm(500, mean = 10, sd = 2)), 
           salary = rnorm(500, mean = 45000, sd = 10000),
           last_review = sample(c("Developing", "Valued", "Top Performer"), 500, replace = TRUE), 
           current_employee_NY = sample(0:1, 500, replace = TRUE)) %>% 
  summary()
```

We might actually be able to do something with this data, but with some limitations. The current_employee_NY variable might give us an idea about turnover, but it could also be filled with people who have retired or have been laid off. As that variable stands, there is just not enough information for us to know the reason why the person is no longer employed.


What questions about turnover could be answered with this data:

```{r}
data.frame(degree = sample(c("CS", "Management", "Biology"), 500, replace = TRUE), 
           degree_type = sample(c("BS", "Masters", "Doctoral"), 500, replace = TRUE),
           hired_NY = sample(0:1, 500, replace = TRUE)) %>% 
  summary()
```


Even if you have all of the necessary data, you need to make sure that the hypotheses actually follow a logic mechanism (while causality is hard to demonstrate, our hypotheses should at least make causal sense).

Does the following hypothesis make sense? If not, how could it be revised?

> People who use online dating services like Tandem are more likely to engage in extramarital affairs.

This is a classic case of odd causality that would just benefit from a more concrete statement. I might repose this question like, "Married people who have accounts to online data services are more likely to engage in extramarital affairs than married people who do not have accounts to online data services."

What about this one?

> Gambling causes people to have higher levels of risk taking behaviors.

Here is some reversed causality. It should probably be, "People who demonstrate higher levels of risk taking behaviors will likely gamble more than people who do not demonstrate risk taking behaviors."

And this one?

> Having more muscle mass leads to greater expenditures on protein powder.

This is a bit of a "chicken or egg" problem. While it is clear that people who have more muscle mass likely spend more money on protein supplements, the question is a hard one to answer. A better question might be, "Do people who engage in anerobic exercise (defined by hours per week) spend more money than people who engage in aerobic exercise?"

Finally?

> People who litter less are more responsive to anti-littering ad campaigns.

This one might be impossible to salvage in its current form. Instead, we would really need to think through the design of an experiment here. We would need to pose a question to find if, "Littering rates decreased after exposure to an anti-littering advertisement."