---
title: "Week 1 Practice"
author: "BDS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
```

Data excites people. If you give people data, there is a good chance that they will take it straight to Excel and pivot table away. This is the type of behavior that cause some people to note that human beings are natural scientists -- we are always on the look-out for patterns and relationships. This practice becomes even more rewarding and engaging when that is about people, from people, and what those people do. The thought is typically, "Hey! I'm a person, so I probably know about what other people do!" 

When people have this data and access to someone with the chops to answer questions, their brains will start creating a glut of questions. As a practicioner of data science, you need to be a guiding light for asking the correct questions. Furthermore, you must remember that great magic happens in the blackbox, but people often like to have explainable answers to the question.

You have the power to get the answers, but you need to understand the question. Not only do you need to understand the question, but sometime you will need to shape the question

Given the following data and question, provide some advice on what questions could actually be answered:

> Do tobacco users utilize riskier betting strategies than non-users?

```{r}
data.frame(cigs_smoked_daily = rnorm(50, 25, 2), 
           betting_strategy = sample(c("conservative", "balanced", "aggressive"), 50, replace = TRUE)) %>% 
  summary()
```

Given the provided data, the proposed question cannot really be answered. While our betting strategy variable is suitable, we would need a binary variable for tobacco use to really answer the question. What we could test, though, is if there are any differences between the betting strategy groups and the number of cigarettes smoked per day. 


> Will more experienced video game players use fewer pulse blasts (i.e., weak blasts, but fast follow-up shots) than players with less experience?

```{r}
data.frame(charge_blast_count = rnorm(50, 75, 15), 
           hours_in_game = rnorm(50, 50, 10)) %>% 
  summary()
```

This data almost seems like it would get at the question, but we need to take just a step back and think about what was asked. The charge_blast_count variable does not capture anything about pulse blasts -- if we have a variable about pulse blasts, we could figure out the rate at which pulse blasts were used. Futhermore, the hours_in_game variable is only telling us how much time someone has spent playing the particular game, not how experienced they are with video games in general. The only question that we can really answer here is if people with more hours in the game use more charge blasts.

> Are there any group differences in page interaction (click rates, scroll rates, etc.) between people using different web browsers?

```{r}
data.frame(seconds_on_page = rnorm(5000, 60, 10), 
           browser = sample(c("Safari", "Chrome", "Firefox", "Edge"), 5000, replace = TRUE)) %>% 
  summary()
```

The browser variable works for us, but the seconds_on_page variable really does not give us much in the way of interaction. Ideally, we would have variables about the number of mouse clicks, the amount of scrolling, and any keystrokes. To that end, all we can really do is tell if there is any difference between the browser groups in the time spent on page.