---
title: "Week 5 Practice"
author: "BDS"
date: "July 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(mirt)

library(dplyr)

```

We have the same json file that we had last week, so nothing has changed there.


```{r}
week5 = jsonlite::read_json("C://Users/berry2006/Documents/projects/courses/bds/exerciseData/week04and05.json", 
                            simplifyVector = TRUE)
```

As with everything, we need to think through some issues before we just dive right into coding the model up. 

What was the nature of our response -- dichotomous or polytomous? Once we know, we can make the first decision. 

```{r}
summary(week5)
```

These are clearly polytomous items, so we are going to be working with a polytomous model. Now, we need to think through our parameters. 

```{r}
week5IRT = week5 %>% 
  select(-id) %>% 
  mirt(data = ., model = 1, 
       itemtype = "graded", verbose = FALSE)

summary(week5IRT)

```

Let's peak at our item trace lines.

```{r}
plot(week5IRT, type = "trace")
```

We clearly have some items that are not providing much information! Some items, however, have response options that appear to discriminate pretty well over the range of $\theta$.

Remember, though, that our factor analysis from last week yielded a pretty good solution with two factors. We should probably take that into account here.

While we did not get much into the exploratory/confirmatory distinction, we can do both here.

Here, we are just specifying two factors and our model will do the rest (this is exploratory).
```{r}
week5IRT2Factor = week5 %>% 
  select(-id) %>% 
  mirt(data = ., model = 2, 
       itemtype = "graded", verbose = FALSE)

summary(week5IRT2Factor, rotate = "promax")
```

Our first factor looks pretty good, but the second factor is a bit different.

Like our factor analysis, we also want to 

If we want to go the confirmatory route, we would do something like the following:

```{r}
confirmatoryModel = mirt.model('
   F1 = 1,3,4,5
   F2 = 2,6,7')

week5IRTConfirmatory = week5 %>% 
  select(-id) %>% 
  mirt(data = ., model = confirmatoryModel, 
       itemtype = "graded", verbose = FALSE)
```

When we look at our loadings, we can see that they are pretty strong -- we would likely have a good chance at accepting this as a confirmatory model.

Let's also compare our single factor model to our exploratory two factor model.

```{r}
anova(week5IRT, week5IRT2Factor)
```

Given the fits, the two factor model is definitely better.