---
title: "Week 13 Practice"
author: "BDS"
date: "July 6, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(dplyr)

library(tidyr)

library(tidytext)

library(sentimentr)
```


We have another RData file to load in:

```{r}
load("C:/Users/sberry5/Documents/project/courses/bds/exerciseData/week13.RData")
```

Nothing too wild with our structure, but we will want to keep an eye on that list column!

```{r}
glimpse(week13)
```

Nothing too wild with our structure, but we will want to keep an eye on that list column!

Let's start really simple and use some of the tidytext stuff. We need to do some prep first -- it was not too long before that list column was going to give us some issues!

```{r}
week13Tidy = week13 %>% 
  select(id, manager, describeManager) %>% 
  ungroup() %>%
  mutate(describeManager = unlist(.$describeManager), 
         describeManager = gsub("^c\\(|\\)$", "", .$describeManager)) 
```

Given a few of our variables, we can do some interesting things. First, we can get a sentiment by id (i.e., the person who offered the manager descriptions):

```{r}
employeeSentiment = week13Tidy %>%
  unnest_tokens(tbl = ., output = word, input = describeManager) %>% 
  group_by(id) %>% 
  inner_join(get_sentiments("bing")) %>% 
  count(sentiment) %>% 
  spread(sentiment, n, fill = 0) %>% 
  mutate(sentiment = positive - negative)

employeeSentiment
```

Employee A192M really did not like his/her manager!

Now, let's look at it from the manager perspective. While they did not rate theirselves, it will give us an idea about the mean sentiment towards the manager:

```{r}
managerSentiment = week13Tidy %>%
  unnest_tokens(tbl = ., output = word, input = describeManager) %>% 
  group_by(manager, id) %>% 
  inner_join(get_sentiments("bing")) %>% 
  count(sentiment) %>% 
  spread(sentiment, n, fill = 0) %>% 
  group_by(manager) %>% 
  summarize(meanNegative = mean(negative), 
            meanPositive = mean(positive), 
            sentiment = meanPositive - meanNegative)

managerSentiment
```

These give us a pretty good idea about the general sentiment directed towards the manager.

At the end of the day, these are purely count based metrics and we have no idea the true strength of any person's words.

For example, lets take a look at a few words offered to our highest and lowest manager:

```{r}
worstManager = managerSentiment$manager[which.min(managerSentiment$sentiment)]

bestManager = managerSentiment$manager[which.max(managerSentiment$sentiment)]

week13$describeManager[week13$manager == worstManager][sample(1:5, 1)]

week13$describeManager[week13$manager == bestManager][sample(1:5, 1)]
```

Striking, right!?! I would say that those words might have some different valences.

Let's try something a bit more intelligent!

```{r}
vadarSent = sentiment(week13Tidy$describeManager, polarity_dt = lexicon::hash_sentiment_vadar)
```

If we want to get an idea about a manager's overall sentiment, we just need to do a little bit of joining!

```{r}
week13Tidy %>% 
  mutate(element_id = 1:nrow(.)) %>% 
  left_join(., vadarSent, by = "element_id") %>% 
  group_by(manager) %>% 
  summarize(meanSent = mean(sentiment))
```

While that one person is still clearly horrible, the manager who has the highest sentiment has changed.