---
title: 'Homework #1'
author: "Your Name Here"
date: "11/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Bronze

Each of the following sections is worth 10 points apiece.

## A Bank

For this example, assume that one teller has an average working time of 3 minutes with a standard deviation of .5 minutes. Also assume that 1 customer comes every 2 minutes.

What are the average waiting times in systems with 1 teller, 2 tellers, and 3 tellers? How many tellers would you recommend using based upon your results. 

```{r}
workTimeMean

workTimeSD

numberTellers

arrivalRate

simulationTime

customer <- trajectory("Customer path") %>% 
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("teller") %>% 
  timeout(function() {rnorm(n = 1, mean = workTimeMean, sd = workTimeSD)}) %>% 
  release("teller")

bank <- simmer("bank") %>% 
  add_resource("teller", capacity = numberTeller, queue_size = 8) %>% 
  add_generator("Customer", customer, function() {c(0, rexp(100, arrivalRate), -1)})

run(bank, until = simulationTime)

result = get_mon_arrivals(bank)

finishers <- result[result$finished == TRUE, ]

finishers$waitingTime <- finishers$end_time - finishers$start_time - finishers$activity_time
```

## A Call Center

In a typical 8-hour shift, customers call a help desk at an exponential rate of 1 person every 4 minutes. The average call can be approximated by a normal distribution with a mean of 10 minutes and a standard deviation of 2 minutes. The help desk manager is trying to decide how many customer service (CS) reps to hire. There is a total of 10 lines that can be split between hold lines and CS lines. Starting with 1 CS rep and going up to 5 CS reps, determine how many CS reps need to be hired to achieve a service level of .99 and keep the mean waiting time below 2 minutes.

```{r}
callTimeMean

callTimeSD

arrivalRate

simulationTime

csNumber

holdNumber

caller <- trajectory("caller") %>% 
  set_attribute("start_time", function() {now(callCenter)}) %>% 
  seize("phoneLine") %>% 
  timeout(function() rnorm(1, callTimeMean, callTimeSD)) %>% 
  release("phoneLine")

callCenter <- simmer("callCenter") %>% 
  add_resource("phoneLine", capacity = csNumber, queue_size = holdNumber) %>% 
  add_generator("caller", caller, mon = 2, function() {c(0, rexp(1000, arrivalRate), -1)})

run(callCenter, 120)

results <- get_mon_arrivals(callCenter)

results$waitingTime <- results$end_time - results$start_time - results$activity_time

sum(results$finished[results$finished == TRUE]) / length(results$finished)

mean(results$waitingTime)
```

# Silver

The following problem is worth 6 points.

## An Airport


```{r}
arrivalRate

inspection1WorkingMean
inspection1WorkingSD

additionalInspection1WorkingMean
additionalInspection1WorkingSD

line2Percentage

simulationTime

passenger <- trajectory("passenger") %>% 
  set_attribute("start_time", function() {now(airport)}) %>% 
  seize("insp1") %>% 
  timeout(function() rnorm(1, inspection1WorkingMean, inspection1WorkingSD)) %>% 
  release("insp1") %>% 
  branch(function() sample(0:1, 1, prob = c(1 - line2Percentage, line2Percentage)) == 1, 
         continue = TRUE,
         trajectory() %>% 
           seize("insp2") %>% 
           timeout(function() rnorm(1, additionalInspection1WorkingMean, additionalInspection1WorkingSD)) %>% 
           release("insp2")
  )
  
airport <- simmer("airport") %>% 
  add_resource("insp1", capacity = 2, queue_size = Inf) %>% 
  add_resource("insp2", capacity = 2, queue_size = Inf) %>% 
  add_generator("passenger", passenger, mon = 2, function() {c(0, rexp(1000, arrivalRate), -1)})

run(airport, simulationTime)
```

# Gold

The following problem is worth 4 points

## An Airport -- Expanded

