---
title: "Lab 1"
author: "Your Name Here"
date: "11/6/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(simmer)

library(simmer.plot)
```

## The Situation

A hospital has a lot of different things going on with it, but we are going to model a pretty easy process. 

Let's model some exciting times: between 12:00am and 8:00am. Given that this is prime emergency room action, we know that people come in about every 15 minutes (according to an exponential distribution. When a patient walks into a hospital (or an emergency room), they are going to visit a registration desk.

From the desk, 75% of patients go directly to x-ray and 25% go to the lab. Once all of those patients are done, they visit the doctor.

The working times are listed below:

Admit: normal(3, .1)

X-ray: normal(15, 3)

Lab: normal(15, 6)

Doctor: normal(30, 6)

You can assume that each of these servers/workstations can only serve one person at a time.

## Process Flow Map

```{r}
library(DiagrammeR)

grViz("
digraph {
  graph [overlap = true, fontsize = 10, rankdir = LR]
  
  node [shape = box, style = filled, color = black, fillcolor = aliceblue]
  A [label = '']
  B [label = '']
  C [label = '']
  D [label = '']

  A->B A->C B->D C->D
}
")
```

## The Model

We are going to see some things that we have not used before: branches and logs.

A branch will let you define subtrajectories and logs will print messages out to the console (they are good for checking your sanity).

```{r}
patient <- trajectory("patient") %>% 
  set_attribute("start_time", function() now(hospital)) %>% 
  seize("desk") %>% 
  log_("I'm at the desk") %>% 
  timeout(function() rnorm(1, 3, .1)) %>% 
  release("desk") %>%
  branch(function() sample(1:2, 1, prob = c(.75, .25)), 
         continue = c(TRUE, TRUE), 
         trajectory() %>%
           log_("In xray") %>% 
           seize("xray") %>% 
           timeout(function() rnorm(1, 15, 6)) %>% 
           release("xray") %>% 
           log_("leaving xray"), 
         trajectory() %>%
           log_("In lab") %>% 
           seize("lab") %>% 
           timeout(function() rnorm(1, 15, 6)) %>% 
           release("lab") %>% 
           log_("leaving lab")) %>% 
  seize("doctor") %>% 
  log_("With doctor") %>% 
  timeout(function() rnorm(1, 30, 6)) %>% 
  release("doctor") %>% 
  log_("leaving doctor")

# If you remove the log messages, you can plot the trajectory.
# This serves as the process flow map.
# plot(patient)

hospital <- simmer("hospital") %>% 
  add_resource("desk", capacity = 1) %>% 
  add_resource("xray", capacity = 1) %>% 
  add_resource("lab", capacity = 1) %>% 
  add_resource("doctor", capacity = 1) %>% 
  add_generator("patient", patient, mon = 2, function() c(0, rexp(100, 1/15), -1))

run(hospital, 480)

results <- get_mon_arrivals(hospital)

resourceResults = get_mon_resources(hospital)

plot(resourceResults, "utilization")

```


After running your initial simulation, what do you see as the bottleneck (the server that is slowing down progress) in this particular system? Make a recommendation for changing this system and implement that change. How many more people does this system get through with your change?

Time permitting consider an option in which nearly everything remains the same, except for the inclusion of another server: a PA. After finishing with the x-ray or lab, patients can see either the doctor (25%) or the PA (75%).