---
title: "Lab 1"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## The Data

Let's play around with some data about apps in the Play Store.

```{r}
playStore = read.csv("http://www.nd.edu/~sberry5/data/googleplaystore.csv")
```

As we look at our data, we see some variables are clearly categorical:

```{r}
str(playStore)
```

Variables like `Category`, `Type`, `Content.Rating`, and potentially `Installs` could be interesting to mess around with from a categorical data analysis perspective. But first...we should probably check to see what is going on with those variables:

```{r}
summary(playStore$Category)

summary(playStore$Type)

summary(playStore$Content.Rating)
```

In looking at our summaries, it appears that we might want to do a bit of cleaning before we dive into any analyses. Why? If we don't we are going to have bad problems with unreliable estimates given cell counts.

We can start with removing low count stuff on the `Content.Rating` variable. The following demonstrates once such way of doing it:

```{r, eval = FALSE}
library(dplyr)

playStore = playStore %>% 
  filter(Content.Rating != "" & Content.Rating != "Unrated" & Content.Rating != "Adults only 18+") %>% 
  mutate(Content.Rating = droplevels(Content.Rating))

playStore = playStore %>%
  filter(Content.Rating != "" & Content.Rating != "Unrated" & Content.Rating != "Adults only 18+") %>%
  filter(Type == "Free" | Type == "Paid") %>% 
  mutate(Content.Rating = droplevels(Content.Rating), 
         Type = droplevels(Type))
```

You will also want to clean up the `Type` variable, but you will likely want to use the "Or" operator (`|`) to create your statement. While there is clearly a bad category in the `Category` variable, it is completely up to you about how you might want to proceed with anything else. 

## Visualize

For the sake of starting out, do a little bit of visualization among some of your categorical variables. For example, producing a visualization for `Content.Rating` and `Type` can help shed some insight into the relationships (or lack thereof):

```{r, eval = FALSE}
library(ggmosaic)

library(dplyr)

playStore %>% 
  select(Content.Rating, Type) %>% 
  table() %>%
  as.data.frame() %>% 
  ggplot() +
  geom_mosaic(aes(weight = Freq, x = product(Content.Rating), fill = Type)) + 
  scale_fill_brewer(type = "qual")
```

Feel free to tackle any additional visualizations that might be of interest to you.

## Analyze

After conducting your visualizations, you might have a pretty good idea about what is going on between your categories. We can use an $\chi^2$ to test those categories for independence.

```{r, eval = FALSE}
chisq.test(playStore$Content.Rating, playStore$Type, correct = FALSE)
```

What does that test result mean in terms of the association between those variables?

It is also worth trying to visualize the directional association between those variables.

```{r, eval = FALSE}
playStore %>% 
  select(Content.Rating, Type) %>% 
  assoc(., shade = TRUE)
```