---
title: "Lab 2"
author: "Your Name Here"
date: "11/13/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Data

Continue using the annual exective compensation file:

```{r}
link <- "http://www3.nd.edu/~sberry5/data/anncomp.csv"
```

As noted in the lecture notes, this data needs some cleaning. You will notice that people are listed multiple times within this data, since it is yearly data -- we are just going to deal with that as it is right now. While there are better ways to handle this, it is a lecture for another time. Another issue is that we might not have data from the same year for everyone. Again, multiple ways of dealing with it -- we could convert all of the nominal dollars to real dollars (using the Consumer Price Index). This would be the best way to deal with that issue.

```{r}
library(rvest)

cpiTable <- read_html("https://www.usinflationcalculator.com/inflation/consumer-price-index-and-annual-percent-changes-from-1913-to-2008/") %>% 
  html_table(header = TRUE) %>% 
  `[[`(1)
```

After a little row/name clean-up, we can join the cpiTable into our compensation data and create an inflation-adjusted total salary variable.

The formula to convert nominal dollars to real dollars is $\frac{\text{nominal}}{\text{price index} / 100}$


## Standard Analyses

In the class notes, we looked at the *SALARY* variable; now let's shift our focus to the *TOTAL_CURR* variable. This variable includes the total for all forms of compensation (salary, stocks, etc.). Let's also add another predictor into the model: *GENDER*. 

### Model 1

First construct a standard model, with *AGE* and *GENDER* as individual predictors. Briefly describe your results.

### Model 2

Create a model with an interaction term for *AGE* and *GENDER*. Discuss and visualize the effects.

## Robust Analyses

You will likely find some problems with your model (e.g., outliers and/or standard error issues). 